% ══════════════════════════════════════════════════════════════════
\chapter{Mathematical Framework and Quantum Algorithms}
\label{ch:framework}

This chapter develops the mathematical machinery used by three quantum
algorithms for charmonium spectroscopy that we reproduce in this thesis:
the Variational Quantum Eigensolver of Gallimore and
Liao~\cite{gallimore2023}, the Variational Imaginary-Time Evolution
algorithm applied to charmonium by Woloshyn~\cite{woloshyn2023}, and the
Trotter--Taylor Imaginary-Time Evolution scheme of Yi et
al.~\cite{yi2025}. Each method takes a different approach to estimating
the eigenvalues of a Hamiltonian matrix on a quantum processor, and each
makes a different choice of qubit encoding for that matrix. We therefore
organise the chapter so that the material common to the three methods is
presented up front (\cref{sec:common}); each method then receives a
dedicated section that develops it in self-contained form, including the
specific encoding it uses, its ansatz (if any), its excited-state
strategy, and any auxiliary topics treated by the source paper.
The Gallimore--Liao section (\cref{sec:gallimore}) presents VQE together
with the Jordan--Wigner encoding and an explicit extension of their
construction to the spin-dependent $N=4$ case. The Woloshyn section
(\cref{sec:woloshyn}) covers VQITE in the direct encoding and includes
the M1 and E1 transition amplitudes that Woloshyn also derives. The
Yi et al.\ section (\cref{sec:yi}) reproduces TTITE, applied to the same
channel Hamiltonians used by the previous two methods. A side-by-side
summary closes the chapter (\cref{sec:framework-summary}).


% ──────────────────────────────────────────────────────────────────
\section{Common Framework}
\label{sec:common}

Before introducing the three algorithms, we collect the material they all
build on: the charmonium Hamiltonian in a finite harmonic-oscillator
basis, the three $4\times 4$ channel matrices, and the general
prescription for expressing a Hermitian matrix as a sum of Pauli strings
on a quantum register.

\subsection{The non-relativistic quark model}
\label{ssec:quark_model}

A charmonium meson is a bound state of a charm quark and its
antiquark~\cite{eichten1978,brambilla2011}. Because the charm mass
$m_c \simeq 1.48~\mathrm{GeV}/c^2$ is large compared with the binding energy,
the internal motion is well described non-relativistically: the relative
coordinate $\mathbf{r} = \mathbf{r}_c - \mathbf{r}_{\bar c}$ satisfies the
two-body Schr\"{o}dinger equation
\begin{equation}
  \label{eq:two_body}
  H\,\psi(\mathbf{r}) =
    \left[-\frac{\nabla^{2}}{2\mu} + V(r)\right]\psi(\mathbf{r})
    = E\,\psi(\mathbf{r}),
\end{equation}
with reduced mass $\mu = m_c/2$. The interaction $V(r)$ has two pieces.
The first is the spin-independent Cornell potential
\begin{equation}
  \label{eq:cornell_potential}
  V_0(r) = -\frac{a}{r} + b\,r,
  \qquad a = \frac{4\alpha_s}{3},
\end{equation}
where $-a/r$ is the colour-Coulomb attraction from one-gluon exchange and
$b\,r$ is the linearly rising flux-tube term that confines the
quarks~\cite{eichten1978}. The second piece is the leading spin-dependent
correction. From the non-relativistic reduction of the Breit interaction,
the dominant term at one-gluon-exchange order is the contact
(hyperfine) interaction~\cite{barnes2005,brambilla2011}
\begin{equation}
  \label{eq:hyperfine}
  V_s(r) =
    \frac{32\pi\,\alpha_s}{9\,m_c^{2}}\,
    \delta_\sigma(\mathbf{r})\,
    \mathbf{S}_c\cdot\mathbf{S}_{\bar c},
\end{equation}
where we have replaced the singular Dirac delta with a normalised Gaussian of
width $\sigma$,
\begin{equation}
  \label{eq:gaussian_delta}
  \delta_\sigma(\mathbf{r}) =
    \left(\frac{\sigma}{\sqrt{\pi}}\right)^{3} e^{-\sigma^{2} r^{2}},
\end{equation}
to regularise the contact term inside our finite basis. The dot product of
the two quark spin operators is fixed by the total spin
$\mathbf{S} = \mathbf{S}_c + \mathbf{S}_{\bar c}$ through the identity
$\mathbf{S}^{2} = \mathbf{S}_c^{2} + \mathbf{S}_{\bar c}^{2}
 + 2\,\mathbf{S}_c\cdot\mathbf{S}_{\bar c}$, giving
\begin{equation}
  \label{eq:spin_dot}
  \mathbf{S}_c\cdot\mathbf{S}_{\bar c} =
  \begin{cases}
    -\tfrac{3}{4} & \text{spin singlet } (S=0),\\[4pt]
    +\tfrac{1}{4} & \text{spin triplet } (S=1).
  \end{cases}
\end{equation}
The phenomenological parameters $\alpha_s$, $b$, $m_c$, and $\sigma$ used
throughout this work are those of Barnes, Godfrey, and
Swanson~\cite{barnes2005} and are listed in \cref{tab:cornell_params}.

\begin{table}[htbp]
\centering
\caption{Cornell-potential parameters used in this work, following the
  charmonium fit of Barnes, Godfrey, and Swanson~\cite{barnes2005}. The
  conversion $\hbar c = 197.326$~MeV$\cdot$fm relates inverse-femtometres to
  MeV. The Coulomb coefficient is $a = 4\alpha_s/3$.}
\label{tab:cornell_params}
\begin{tabular}{lll}
\toprule
Parameter & Symbol & Value \\
\midrule
Strong coupling          & $\alpha_s$    & $0.5461$ \\
String tension           & $b$           & $0.1425~\mathrm{GeV}^{2}$ \\
Charm-quark mass         & $m_c$         & $1.4794~\mathrm{GeV}$ \\
Gaussian width           & $\sigma$      & $1.0946~\mathrm{GeV}$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Radial equation and the three channels}
\label{ssec:channels}

Because $V_0(r)$ depends only on $r = |\mathbf{r}|$ and the spin operator in
\cref{eq:hyperfine} commutes with the orbital angular momentum, eigenstates
of $H$ can be classified by the spectroscopic quantum numbers
${}^{2S+1}L_J$. Writing the wavefunction as
\begin{equation}
  \label{eq:separation}
  \psi(\mathbf{r}) = R_{n\ell}(r)\,Y_{\ell m}(\hat{\mathbf{r}})\,\chi_{S M_S},
\end{equation}
and defining the reduced radial function $u_{n\ell}(r) = r R_{n\ell}(r)$, the
Schr\"{o}dinger equation reduces to the one-dimensional eigenvalue problem
\begin{equation}
  \label{eq:radial}
  \left[
    -\frac{1}{2\mu}\frac{\mathrm{d}^{2}}{\mathrm{d}r^{2}}
    + V_0(r)
    + \frac{\ell(\ell+1)}{2\mu r^{2}}
    + \langle\mathbf{S}_c\cdot\mathbf{S}_{\bar c}\rangle\,
      \frac{32\pi\alpha_s}{9 m_c^{2}}\,\delta_\sigma(r)
  \right] u_{n\ell}(r) = E_{n\ell}\,u_{n\ell}(r),
\end{equation}
where the centrifugal barrier $\ell(\ell+1)/(2\mu r^2)$ is the standard
remnant of the angular Laplacian, and the spin expectation value takes the
discrete values of \cref{eq:spin_dot}.

In this work we focus on the three low-lying channels that have served as
benchmarks for quantum simulations of quarkonium~\cite{woloshyn2023}:
\begin{itemize}
  \item ${}^1S_0$ (associated with the $\eta_c$ meson): $\ell=0$, $S=0$. The
    centrifugal barrier vanishes and the contact term contributes with
    $\langle\mathbf{S}_c\cdot\mathbf{S}_{\bar c}\rangle = -3/4$.
  \item ${}^3S_1$ (associated with the $J/\psi$ meson): $\ell=0$, $S=1$. As
    for ${}^1S_0$ but with $\langle\mathbf{S}_c\cdot\mathbf{S}_{\bar c}\rangle
    = +1/4$, producing the celebrated $\eta_c$--$J/\psi$ hyperfine splitting.
  \item ${}^1P_1$ (associated with the $h_c$ meson): $\ell=1$, $S=0$. The
    contact term contributes only at the origin, and because the
    $\ell=1$ wavefunction vanishes there the spin-dependent term is
    numerically negligible for $P$-waves; the centrifugal barrier, however,
    shifts the energies up substantially.
\end{itemize}
Each channel thus generates a different one-dimensional radial Hamiltonian,
which we now discretise.

\subsection{Harmonic-oscillator basis expansion}
\label{ssec:ho_basis}

The radial functions $u_{n\ell}(r)$ admit a convenient expansion in the
eigenstates of a three-dimensional isotropic harmonic oscillator (HO),
\begin{equation}
  \label{eq:ho_states}
  \phi_{n\ell}(r) =
    N_{n\ell}\,r^{\ell+1}\,
    \exp\!\left(-\frac{r^{2}}{2 b^{2}}\right)
    L_{n}^{\ell + 1/2}\!\left(\frac{r^{2}}{b^{2}}\right),
\end{equation}
with normalisation
\(N_{n\ell} = \sqrt{2\,n!\, b^{-(2\ell+3)} /\,\Gamma(n+\ell+3/2)}\)
and $L_n^{\alpha}$ the associated Laguerre polynomials. The length scale
$b = (\mu\omega)^{-1/2}$ is determined by the oscillator frequency $\omega$,
which is chosen variationally so that the ground state of the basis already
captures most of the physical ground-state wavefunction~\cite{gallimore2023}.

We truncate the basis to the first $N=4$ states for each channel,
\begin{equation}
  \label{eq:basis_expansion}
  u_{n\ell}(r) \;\approx\; \sum_{k=0}^{N-1} c_k^{(n)}\,\phi_{k\ell}(r),
  \qquad N = 4,
\end{equation}
which turns \cref{eq:radial} into the finite-dimensional eigenvalue problem
\begin{equation}
  \label{eq:matrix_problem}
  \sum_{k=0}^{N-1} H_{mk}^{(\ell,S)}\,c_k^{(n)} = E_n\,c_m^{(n)},
\end{equation}
where the channel-dependent matrix elements are
\begin{equation}
  \label{eq:matrix_elements}
  H_{mk}^{(\ell,S)} =
    \langle\phi_{m\ell}|\,T\,|\phi_{k\ell}\rangle
    + \langle\phi_{m\ell}|\,V_0\,|\phi_{k\ell}\rangle
    + \langle\mathbf{S}_c\cdot\mathbf{S}_{\bar c}\rangle\,
      \frac{32\pi\alpha_s}{9 m_c^{2}}\,
      \langle\phi_{m\ell}|\,\delta_\sigma\,|\phi_{k\ell}\rangle.
\end{equation}
Here $T$ denotes the kinetic term plus, for $\ell>0$, the centrifugal barrier.

\begin{remark}[Hylleraas--Undheim--MacDonald theorem]
The truncation in \cref{eq:basis_expansion} introduces a controlled
approximation: the eigenvalues of the truncated matrix
$H^{(\ell,S)} \in \mathbb{R}^{N\times N}$ are guaranteed to be upper
bounds on the corresponding exact eigenvalues, and they are monotonically
non-increasing in $N$~\cite{macdonald1933}. Increasing $N$ therefore
always improves the variational estimate; $N=4$ provides a useful balance
between physical accuracy and the qubit cost of the encodings used by the
three methods that follow.
\end{remark}

\subsection{Matrix elements}
\label{ssec:matrix_elements}

The matrix elements in \cref{eq:matrix_elements} can be evaluated in closed
form using standard identities for the Laguerre polynomials and the
hypergeometric function. We collect the key results here; complete derivations
follow the methodology of Gallimore and Liao~\cite{gallimore2023}.

\paragraph{Kinetic energy.}
The kinetic operator, expressed through the ladder operators of the
oscillator, is tridiagonal in the HO basis. For $\ell=0$ one finds
\begin{equation}
  \label{eq:kinetic_matrix}
  \langle m | T | k \rangle = \frac{\omega}{2}\Bigl[
    \bigl(2k + \tfrac{3}{2}\bigr)\delta_{mk}
    -\sqrt{k(k+\tfrac{1}{2})}\,\delta_{m,k-1}
    -\sqrt{(k+1)(k+\tfrac{3}{2})}\,\delta_{m,k+1}
  \Bigr].
\end{equation}
For $\ell>0$ the centrifugal term is absorbed into a shifted diagonal,
$2k + \ell + 3/2$, with the off-diagonal couplings inheriting the same
square-root structure with the index $\ell + 1/2$ in place of $1/2$.

\paragraph{Coulomb term.}
The matrix element of $1/r$ can be expressed in terms of a generalised
hypergeometric function ${}_3F_2$~\cite{gallimore2023}:
\begin{equation}
  \label{eq:coulomb_matrix}
  \langle m | r^{-1} | k \rangle =
    \frac{(-1)^{m+k}\,4 b^{-1}}{\pi (1 + 2k)}
    \sqrt{\frac{\Gamma(m+\tfrac{3}{2})\,\Gamma(k+\tfrac{3}{2})}{m!\,k!}}\,
    {}_3F_2\!\left(\tfrac{1}{2},1,-m;\;\tfrac{3}{2},\tfrac{1}{2}-k;\;1\right).
\end{equation}

\paragraph{Linear-confinement term.}
The matrix element of $r$ involves the Gauss hypergeometric ${}_2F_1$:
\begin{equation}
  \label{eq:linear_matrix}
  \langle m | r | k \rangle =
    \frac{(-1)^{m+k}\,4 b}{\pi (1 - 4k^{2})}
    \sqrt{\frac{\Gamma(m+\tfrac{3}{2})\,\Gamma(k+\tfrac{3}{2})}{m!\,k!}}\,
    {}_2F_1\!\left(2,-m;\;\tfrac{3}{2}-k;\;1\right).
\end{equation}

\paragraph{Contact term.}
For the regularised contact interaction with the Gaussian of
\cref{eq:gaussian_delta}, the matrix element reduces to a single Gaussian
integral over a polynomial in $r^{2}$, which is again closed-form:
\begin{equation}
  \label{eq:contact_matrix}
  \langle\phi_{m\ell}|\,\delta_\sigma\,|\phi_{k\ell}\rangle
  = \left(\frac{\sigma}{\sqrt{\pi}}\right)^{3}\!\!
    \int_0^{\infty}\!\!\phi_{m\ell}(r)\,e^{-\sigma^{2} r^{2}}\,
    \phi_{k\ell}(r)\,\mathrm{d}r.
\end{equation}
For $\ell>0$ the factor $r^{2\ell+2}$ in the integrand suppresses the
contribution near the origin; in the limit $\sigma\to\infty$ the integral
vanishes for $\ell>0$ and reproduces $|\phi_{m\ell}(0)\phi_{k\ell}(0)|$
for $\ell=0$, recovering the unregularised Dirac-delta result.

\begin{remark}
The signs $(-1)^{m+k}$ in \cref{eq:coulomb_matrix,eq:linear_matrix} encode
the alternating parity of the Laguerre polynomials at the origin and are
essential for obtaining the correct off-diagonal Cornell-potential matrix
elements. In practice all of these closed-form expressions are evaluated to
high precision via the \texttt{mpmath} library and verified against direct
numerical quadrature.
\end{remark}

\subsection{The three channel Hamiltonian matrices}
\label{ssec:matrices}

Assembling \cref{eq:matrix_elements} channel by channel with the parameters
of \cref{tab:cornell_params} and $\omega = 1.2~\mathrm{fm}^{-1}$ produces the
following three $4\times 4$ matrices (in units of $\mathrm{fm}^{-1}$, with
conversion to MeV via $\hbar c = 197.326~\mathrm{MeV}\!\cdot\!\mathrm{fm}$):
\begin{align}
H^{({}^1S_0)} &= \begin{pmatrix}
   0.9431 & -0.8733 & -0.7690 & -0.5601\\
  -0.8733 &  3.3365 & -0.5646 & -0.8648\\
  -0.7690 & -0.5646 &  5.4382 & -0.1566\\
  -0.5601 & -0.8648 & -0.1566 &  7.3451
\end{pmatrix},
\label{eq:H_1S0}\\[4pt]
H^{({}^3S_1)} &= \begin{pmatrix}
   1.0946 & -0.7114 & -0.6111 & -0.4112\\
  -0.7114 &  3.5406 & -0.3910 & -0.6989\\
  -0.6111 & -0.3910 &  5.6122 & -0.0119\\
  -0.4112 & -0.6989 & -0.0119 &  7.5104
\end{pmatrix},
\label{eq:H_3S1}\\[4pt]
H^{({}^1P_1)} &= \begin{pmatrix}
   2.8561 & -0.2395 & -0.3827 & -0.2282\\
  -0.2395 &  4.9190 & -0.0373 & -0.5097\\
  -0.3827 & -0.0373 &  6.8114 &  0.4058\\
  -0.2282 & -0.5097 &  0.4058 &  8.6034
\end{pmatrix}.
\label{eq:H_1P1}
\end{align}
Each matrix is real and symmetric, a consequence of the Hermiticity of the
underlying operator and the choice of real-valued HO basis functions. The
${}^1P_1$ matrix lies higher in absolute energy because of the
$\ell=1$ centrifugal contribution to the diagonal; the spin-dependent shift
between ${}^1S_0$ and ${}^3S_1$ is visible already in the difference between
their diagonal entries.

\begin{remark}[Errata in the literature]
We note that in the published matrices for ${}^1P_1$~\cite{woloshyn2023} the
$(3,3)$ element was reported as $7.5104$, coinciding numerically with the
$(3,3)$ entry of $H^{({}^3S_1)}$. Cross-checking against the variational
eigenvalues and the diagonal pattern of the kinetic + centrifugal +
confinement contributions identifies this as a copy-paste error; the correct
value $8.6034$ is used in \cref{eq:H_1P1}.
\end{remark}

Diagonalising each $H^{(\ell,S)}$ by classical means immediately yields the
exact (within the truncated basis) energy spectrum of its channel. Doing so on
a quantum computer, however, requires expressing the operator as a sum of
Pauli strings acting on a register of qubits.

\subsection{Pauli decomposition of a Hermitian matrix}
\label{ssec:pauli_decomp}

A common building block for all three algorithms is the expansion of the
channel Hamiltonian in the basis of Pauli strings. Any Hermitian operator
$H$ on $n$ qubits can be written as a real linear combination of the
$4^{n}$ Pauli strings
$\{P_k\} \subset \{I,X,Y,Z\}^{\otimes n}$,
\begin{equation}
  \label{eq:pauli_sum}
  H = \sum_{k=1}^{4^{n}} c_k\,P_k,
  \qquad c_k = \frac{1}{2^{n}}\,\Tr(H\,P_k)\in\mathbb{R}.
\end{equation}
The relation \cref{eq:pauli_sum} follows from the orthogonality of Pauli
strings under the Hilbert--Schmidt inner product,
$\Tr(P_j P_k) = 2^{n}\delta_{jk}$. In practice only a small subset of the
$4^{n}$ coefficients are nonzero; for real symmetric Hamiltonians, all
strings containing an odd number of $Y$ factors vanish identically, leaving
$3^{n}$ candidate terms.

Once $H$ has been put in the form \cref{eq:pauli_sum}, each Pauli string can
be implemented as a basis rotation on the quantum register followed by a
diagonal measurement, and the expectation value $\langle H\rangle$ recovered
as the weighted sum $\sum_k c_k\langle P_k\rangle$. The remaining question
is which qubit register to use---i.e., how to associate physical basis
states with bit-strings. The three source papers make two different
choices, and we present each choice within the section that uses it.


% ──────────────────────────────────────────────────────────────────
\section{Gallimore \& Liao --- Variational Quantum Eigensolver}
\label{sec:gallimore}

Gallimore and Liao~\cite{gallimore2023} apply the Variational Quantum
Eigensolver (VQE)~\cite{peruzzo2014,mcclean2016} to a spinless
Cornell-potential charmonium Hamiltonian truncated to $N=3$
harmonic-oscillator basis states. Their construction maps the
second-quantised one-body operator to qubits via the Jordan--Wigner
transformation---exposing the problem on a register of three qubits with
a one-particle constraint---and uses a two-parameter Unitary Coupled
Cluster ansatz tailored to that constraint. In this section we reproduce
their development, pointing out the parts that generalise without
modification to the spin-dependent $N=4$ matrices of \cref{sec:common}.
We close (\cref{ssec:gallimore_extension}) with the explicit extension
required to apply VQE on equal footing with the other two methods later
in the thesis.

\subsection{The variational principle}
\label{ssec:vqe_principle}

For any Hermitian Hamiltonian $H$ on a finite Hilbert space, every
normalised state $\ket{\psi}$ obeys
\begin{equation}
  \label{eq:rayleigh_ritz}
  E_0 \;\leq\; \bra{\psi}H\ket{\psi},
\end{equation}
where $E_0$ is the smallest eigenvalue of $H$, with equality if and only if
$\ket{\psi}$ lies in the corresponding eigenspace. Restricting the search to
a family of \emph{trial states} $\{\ket{\psi(\bm{\theta})}\}$ parameterised by a
classical vector $\bm{\theta}\in\mathbb{R}^{p}$ gives the upper bound
\begin{equation}
  \label{eq:vqe_cost}
  E_0 \;\leq\;
  \min_{\bm{\theta}} \;
  E(\bm{\theta}),
  \qquad
  E(\bm{\theta}) \equiv \bra{\psi(\bm{\theta})} H \ket{\psi(\bm{\theta})}.
\end{equation}
The VQE algorithm computes $E(\bm{\theta})$ on the quantum processor and
minimises it on a classical computer; the resulting parameters
$\bm{\theta}^{\star}$ characterise an approximation $\ket{\psi(\bm{\theta}^{\star})}$
to the ground state. The accuracy of this approximation is controlled by
the \emph{expressibility} of the ansatz: if the parameterised manifold
$\{\ket{\psi(\bm{\theta})}\}$ contains the true ground state then the
minimum of \cref{eq:vqe_cost} coincides with $E_0$; otherwise it provides
a (variational) upper bound.

The remaining ingredients are (i) a qubit encoding of $H$, (ii) a choice
of ansatz, (iii) a procedure for measuring $E(\bm{\theta})$ from circuit
shots, and (iv) a classical optimiser for $\bm{\theta}$. The next four
subsections cover them in turn.

\subsection{Jordan--Wigner encoding}
\label{ssec:jw_encoding}

The encoding chosen by Gallimore and Liao originates in quantum chemistry.
We treat the HO basis functions as orbitals occupied by a single
charm--anticharm pair and introduce a fermionic creation operator
$a_n^{\dagger}$ for each orbital, with the canonical anticommutation
relations
\begin{equation}
  \label{eq:fermion_anticom}
  \{a_m, a_n^{\dagger}\} = \delta_{mn},
  \qquad
  \{a_m, a_n\} = \{a_m^{\dagger}, a_n^{\dagger}\} = 0.
\end{equation}
The channel Hamiltonian is then the second-quantised one-body operator
\begin{equation}
  \label{eq:second_quantised}
  \widehat{H}^{(\ell,S)} =
    \sum_{m,n=0}^{N-1} H_{mn}^{(\ell,S)}\;a_m^{\dagger} a_n,
\end{equation}
acting on the Fock space of $N$ orbitals; physical states are those
containing exactly one excitation, i.e.\ the $N$ states
$\{a_n^{\dagger}|\mathrm{vac}\rangle\}_{n=0}^{N-1}$. Restricting to this
single-particle sector reproduces the original $N\times N$ matrix problem.

\paragraph{The Jordan--Wigner transformation.}
The Jordan--Wigner (JW) transformation~\cite{nielsen2010} maps each orbital
to a dedicated qubit and faithfully reproduces the fermionic algebra
\cref{eq:fermion_anticom} in terms of Pauli operators,
\begin{align}
  \label{eq:jw_a_dagger}
  a_n^{\dagger} &=
    \frac{1}{2}\,\Bigl(\,\prod_{j<n} Z_j\Bigr)\,(X_n - i\,Y_n),\\[3pt]
  \label{eq:jw_a}
  a_n &=
    \frac{1}{2}\,\Bigl(\,\prod_{j<n} Z_j\Bigr)\,(X_n + i\,Y_n).
\end{align}
The trailing string of $Z$ operators on lower-indexed qubits encodes the
fermionic sign that arises when a creation operator anticommutes past
already-occupied orbitals. With this mapping, $N$ orbitals require $N$
qubits, of which only $\binom{N}{1}=N$ computational-basis states are
physical (for $N=4$, these are
$|0001\rangle,|0010\rangle,|0100\rangle,|1000\rangle$).

\paragraph{Pauli forms of the one-body operators.}
Substituting \cref{eq:jw_a_dagger,eq:jw_a} into the bilinears
$a_m^{\dagger} a_n$ yields concrete Pauli strings. The number operator at
orbital $n$ is diagonal,
\begin{equation}
  \label{eq:jw_number}
  a_n^{\dagger} a_n = \frac{1}{2}(I - Z_n).
\end{equation}
Adjacent hopping ($|m-n|=1$, e.g.\ between orbitals $n$ and $n+1$) gives the
familiar $XX{+}YY$ pattern,
\begin{equation}
  \label{eq:jw_adjacent}
  a_n^{\dagger} a_{n+1} + a_{n+1}^{\dagger} a_n
  = \frac{1}{2}\bigl(X_n X_{n+1} + Y_n Y_{n+1}\bigr),
\end{equation}
while non-adjacent hops pick up the JW $Z$-string,
\begin{align}
  \label{eq:jw_skip1}
  a_n^{\dagger} a_{n+2} + a_{n+2}^{\dagger} a_n
    &= \frac{1}{2}\bigl(X_n Z_{n+1} X_{n+2} + Y_n Z_{n+1} Y_{n+2}\bigr),\\
  \label{eq:jw_skip2}
  a_n^{\dagger} a_{n+3} + a_{n+3}^{\dagger} a_n
    &= \frac{1}{2}\bigl(X_n Z_{n+1} Z_{n+2} X_{n+3}
                       + Y_n Z_{n+1} Z_{n+2} Y_{n+3}\bigr).
\end{align}
Combining these expressions with the coefficients $H_{mn}^{(\ell,S)}$
from \cref{ssec:matrices} produces a Pauli-string decomposition of
$\widehat{H}^{(\ell,S)}$.

\paragraph{The $N=4$ decomposition in full.}
For $N=3$ orbitals (Gallimore and Liao's original setting) the diagonals
contribute one identity and three single-$Z$ terms, the three off-diagonals
contribute pairs of $XX{+}YY$ or $XZX{+}YZY$ strings, giving ten Pauli
terms in total. For the $N=4$ extension used in this thesis, the same
construction is carried out for the four orbitals on four qubits. Denoting
the channel matrix elements $H_{mn} \equiv H_{mn}^{(\ell,S)}$ for brevity,
and using the real symmetry $H_{mn}=H_{nm}$ to combine each off-diagonal
pair $H_{mn}(a_m^{\dagger}a_n + a_n^{\dagger}a_m)$, the explicit Pauli
expansion is
\begin{align}
\widehat{H}^{(\ell,S)}
  &= \underbrace{\tfrac{1}{2}\bigl(H_{00}+H_{11}+H_{22}+H_{33}\bigr)\,I}_{\text{identity}}
  \nonumber\\
  &\quad
  \underbrace{-\tfrac{1}{2}H_{00}\,Z_0
    -\tfrac{1}{2}H_{11}\,Z_1
    -\tfrac{1}{2}H_{22}\,Z_2
    -\tfrac{1}{2}H_{33}\,Z_3}_{\text{4 diagonal }Z\text{ terms}}
  \nonumber\\[2pt]
  &\quad
  \underbrace{+\tfrac{1}{2}H_{01}\bigl(X_0 X_1 + Y_0 Y_1\bigr)
    +\tfrac{1}{2}H_{12}\bigl(X_1 X_2 + Y_1 Y_2\bigr)
    +\tfrac{1}{2}H_{23}\bigl(X_2 X_3 + Y_2 Y_3\bigr)}_{\text{3 adjacent pairs}}
  \nonumber\\[2pt]
  &\quad
  \underbrace{+\tfrac{1}{2}H_{02}\bigl(X_0 Z_1 X_2 + Y_0 Z_1 Y_2\bigr)
    +\tfrac{1}{2}H_{13}\bigl(X_1 Z_2 X_3 + Y_1 Z_2 Y_3\bigr)}_{\text{2 skip-1 pairs}}
  \nonumber\\[2pt]
  &\quad
  \underbrace{+\tfrac{1}{2}H_{03}\bigl(X_0 Z_1 Z_2 X_3
                                       + Y_0 Z_1 Z_2 Y_3\bigr)}_{\text{1 skip-2 pair}}.
  \label{eq:jw_N4_full}
\end{align}
Counting strings gives one identity, four single-$Z$ terms, three pairs of
adjacent $XX{+}YY$ strings ($6$ strings), two pairs of skip-$1$
$XZX{+}YZY$ strings ($4$ strings), and one pair of skip-$2$
$XZZX{+}YZZY$ strings ($2$ strings), totalling
$1 + 4 + 6 + 4 + 2 = 17$ Pauli operators acting on four qubits.

\begin{remark}[Structure of the coefficients]
Each Pauli string in \cref{eq:jw_N4_full} carries a coefficient that is
either a single diagonal entry $H_{nn}$ (for the $Z_n$ terms) or a single
off-diagonal entry $H_{mn}$ (for each member of a pair). Crucially, the
\emph{operator structure} of the Pauli decomposition---which strings
appear, and with what relative signs---is independent of which channel we
work with: the channel-specific information enters only through the
numerical coefficients. The same 17 Pauli strings of
\cref{eq:jw_N4_full} therefore describe $\widehat{H}^{({}^1S_0)}$,
$\widehat{H}^{({}^3S_1)}$, and $\widehat{H}^{({}^1P_1)}$, with the
substitutions $H_{mn} \to H_{mn}^{({}^1S_0)}$, $H_{mn}^{({}^3S_1)}$, and
$H_{mn}^{({}^1P_1)}$ from \cref{eq:H_1S0,eq:H_3S1,eq:H_1P1}.
\end{remark}

\subsection{Unitary Coupled Cluster ansatz}
\label{ssec:ucc_ansatz}

Within the single-particle sector, the trial state must remain
particle-number-conserving. Gallimore and Liao construct an ansatz from
that constraint by exponentiating particle-conserving generators. Starting
from the reference state $\ket{0\cdots 01}$ (one excitation in the lowest
orbital), the Unitary Coupled Cluster (UCC) operator is
\begin{equation}
  \label{eq:ucc_operator}
  U_{\mathrm{UCC}}(\theta_1,\ldots,\theta_{N-1}) =
  \exp\!\left\{
    \sum_{k=1}^{N-1}\theta_k\,(a_k^{\dagger}a_0 - a_0^{\dagger}a_k)
  \right\}.
\end{equation}
Each generator $a_k^{\dagger}a_0 - a_0^{\dagger}a_k$ is anti-Hermitian and
preserves particle number, so $U_{\mathrm{UCC}}$ is unitary and maps
single-particle states to single-particle states. With $N-1$ independent
parameters the operator generates the most general real superposition in
the single-particle sector,
\begin{equation}
  \label{eq:ucc_state}
  \ket{\psi(\bm{\theta})} =
    \cos\theta_{N-1}\,\ket{0\cdots 01}
    + \sin\theta_{N-1}\cos\theta_{N-2}\,\ket{0\cdots 010}
    + \cdots
    + \sin\theta_{N-1}\cdots\sin\theta_1\,\ket{10\cdots 0},
\end{equation}
i.e.\ the $\mathbb{S}^{N-1}$ spherical parameterisation lifted into the
single-particle sector. For $N=3$ this is the two-parameter ansatz of
Gallimore and Liao~\cite{gallimore2023}; for $N=4$ it provides three
parameters and exhausts the real single-particle subspace.

\paragraph{Circuit implementation via Givens rotations.}
A natural circuit-level realisation of \cref{eq:ucc_state} uses a chain of
\emph{Givens rotations} between adjacent orbitals. Within the
single-particle subspace of two adjacent qubits ($j$, $j+1$), a Givens
rotation is the orthogonal $2\times 2$ rotation acting on the basis
$\{\ket{01},\ket{10}\}$:
\begin{equation}
  \label{eq:givens_def}
  G_{j,j+1}(\theta):\quad
  \begin{cases}
    \ket{00} \mapsto \ket{00},\\
    \ket{01} \mapsto \cos\theta\,\ket{01} + \sin\theta\,\ket{10},\\
    \ket{10} \mapsto -\sin\theta\,\ket{01} + \cos\theta\,\ket{10},\\
    \ket{11} \mapsto \ket{11},
  \end{cases}
\end{equation}
which is precisely the exponential
$\exp\bigl\{\theta\,(a_{j+1}^{\dagger}a_j - a_j^{\dagger}a_{j+1})\bigr\}$
restricted to the two qubits in question. A standard hardware decomposition
in terms of one- and two-qubit native gates is
\begin{equation}
  \label{eq:givens_decomp}
  G_{j,j+1}(\theta) =
    \mathrm{CNOT}_{j+1\to j}\;
    \bigl(R_y(2\theta)\bigr)_{j+1}^{\mathrm{ctrl}\,j}\;
    \mathrm{CNOT}_{j+1\to j},
\end{equation}
i.e.\ two CNOTs and one controlled-$R_y$, giving the Givens block constant
depth.

Acting on the reference state $\ket{0\cdots 01}$ (qubit $0$ occupied, all
others empty) with a chain of Givens rotations between adjacent qubits
produces the spherical-coordinate expansion of \cref{eq:ucc_state}
directly:
\begin{equation*}
  \ket{0\cdots 01}
  \;\xrightarrow{G_{0,1}(\theta_{N-1})}\;
  \cdots
  \;\xrightarrow{G_{N-2,N-1}(\theta_1)}\;
  \ket{\psi(\bm{\theta})},
\end{equation*}
with each rotation transferring the particle one orbital further from the
reference. The resulting circuit for $N=4$ is shown in
\cref{fig:ucc_ansatz}; its depth is $N-1$ Givens blocks, growing linearly
in the basis size.

\begin{figure}[htbp]
\centering
\begin{quantikz}
  \lstick{$\ket{0}_0$} & \gate{X} & \gate[2]{G(\theta_3)} & \qw                   & \qw                   & \qw \\
  \lstick{$\ket{0}_1$} & \qw      &                       & \gate[2]{G(\theta_2)} & \qw                   & \qw \\
  \lstick{$\ket{0}_2$} & \qw      & \qw                   &                       & \gate[2]{G(\theta_1)} & \qw \\
  \lstick{$\ket{0}_3$} & \qw      & \qw                   & \qw                   &                       & \qw
\end{quantikz}
\caption{Three-parameter UCC ansatz for the $N=4$ Jordan--Wigner encoding.
  An $X$ gate prepares the reference state $\ket{0001}$ from
  $\ket{0000}$; three Givens rotations
  $G_{j,j+1}(\theta_{N-1-j})$, each decomposed into two CNOTs and a
  controlled-$R_y$ via \cref{eq:givens_decomp}, propagate the single
  excitation through the chain and generate the spherical superposition
  of \cref{eq:ucc_state}. Particle number is exactly conserved by
  construction.}
\label{fig:ucc_ansatz}
\end{figure}

The circuit has $1$ single-qubit $X$ gate, $3$ controlled-$R_y$ gates, and
$6$ CNOTs in total---comparable in depth to the Woloshyn ansatz of
\cref{fig:woloshyn_ansatz} despite using twice as many qubits, because the
single-particle constraint sharply restricts the relevant subspace.

\subsection{Energy measurement on a quantum processor}
\label{ssec:vqe_measurement}

Given the ansatz $\ket{\psi(\bm{\theta})}$ and the Pauli decomposition
$H = \sum_k c_k P_k$ from \cref{eq:pauli_sum}, the energy estimator is
linear in the Pauli expectation values,
\begin{equation}
  \label{eq:vqe_energy_estimator}
  \widehat{E}(\bm{\theta}) = \sum_k c_k\,\langle P_k\rangle_{\bm{\theta}},
\end{equation}
where each $\langle P_k\rangle_{\bm{\theta}}$ is estimated separately by
preparing $\ket{\psi(\bm{\theta})}$ many times and measuring $P_k$.

A measurement in the computational basis directly gives access to
$\langle Z_i\rangle$, but $X$ and $Y$ observables require a basis rotation
applied before the measurement. The rotations are obtained from the
eigenbasis identities $X = H Z H$ and $Y = (S H) Z (H S^{\dagger})$, giving
the standard recipe
\begin{align}
  \label{eq:basis_rotations}
  \langle X_i\rangle &=
    \text{measure }Z_i \text{ after applying } H \text{ on qubit } i,\\[2pt]
  \langle Y_i\rangle &=
    \text{measure }Z_i \text{ after applying } S^{\dagger} H \text{ on qubit } i.
\end{align}
Each Pauli string $P_k$ is mapped to a diagonal observable by applying
these single-qubit basis rotations to every non-identity factor before the
final measurement.

\begin{remark}[Shot noise]
With a finite number $N_{\mathrm{shots}}$ of repetitions per circuit, the
estimator \cref{eq:vqe_energy_estimator} fluctuates around its true value
with a statistical uncertainty
\(
  \sigma_{\widehat{E}}
  = \sqrt{\sum_k c_k^{2}\,\sigma_{P_k}^{2} / N_{\mathrm{shots}}},
\)
where $\sigma_{P_k}^{2} \leq 1$ for any Pauli. The relative error therefore
decreases only as $1/\sqrt{N_{\mathrm{shots}}}$, which makes shot count an
important resource and motivates the choice of optimiser discussed next.
\end{remark}

\subsection{Classical optimisation}
\label{ssec:vqe_optimisation}

The cost function $\widehat{E}(\bm{\theta})$ is, in general, non-convex and
contaminated by shot noise. Two broad families of classical optimisers are
available:

\begin{itemize}\setlength{\itemsep}{2pt}
  \item \textbf{Gradient-free methods} (e.g.\ COBYLA, Nelder--Mead) only
    require evaluations of $\widehat{E}$ and are robust to moderate
    statistical noise. They typically converge in tens to a few hundreds of
    iterations for low-dimensional parameter spaces.
  \item \textbf{Gradient-based methods}, which require partial derivatives
    of $\widehat{E}$ at the cost of additional circuit evaluations. The
    \emph{parameter-shift rule} (\cref{ssec:parameter_shift} below)
    provides exact analytic gradients for rotation gates, but each gradient
    component doubles the number of circuits.
\end{itemize}

Gallimore and Liao use COBYLA, which balances simplicity, noise tolerance,
and convergence speed for low-parameter ansatze. \Cref{fig:vqe_loop_recap}
recalls the resulting hybrid loop already introduced in
\cref{ch:introduction}: the classical optimiser proposes a parameter update,
the quantum processor evaluates $\widehat{E}(\bm{\theta})$, and the cycle
repeats until convergence.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.5cm and 2.4cm,
    box/.style={
      draw, rounded corners=4pt, minimum width=3.4cm,
      minimum height=0.9cm, align=center, font=\small
    },
    arrow/.style={-{Stealth[length=6pt]}, thick},
  ]
  \node[box, fill=blue!8] (prep) {Prepare $\ket{\psi(\bm{\theta})}$};
  \node[box, fill=blue!8, right=of prep] (meas)
    {Measure $\{\langle P_k\rangle\}$};
  \node[box, fill=orange!10, below=of meas] (eval)
    {$\widehat{E}(\bm{\theta}) = \sum_k c_k\langle P_k\rangle$};
  \node[box, fill=orange!10, below=of prep] (opt)
    {Update $\bm{\theta}$ (COBYLA)};
  \draw[arrow] (prep) -- (meas);
  \draw[arrow] (meas) -- (eval);
  \draw[arrow] (eval) -- (opt);
  \draw[arrow] (opt) -- (prep);
  \node[font=\footnotesize\itshape, above=0.15cm of prep, anchor=south west,
        xshift=-0.25cm] {Quantum};
  \node[font=\footnotesize\itshape, below=0.15cm of opt, anchor=north west,
        xshift=-0.25cm] {Classical};
\end{tikzpicture}
\caption{The VQE hybrid loop specialised to charmonium spectroscopy.
  The quantum processor prepares the UCC ansatz of \cref{eq:ucc_state},
  measures the Pauli expectation values in
  \cref{eq:vqe_energy_estimator}, and returns the energy estimate to a
  classical optimiser.}
\label{fig:vqe_loop_recap}
\end{figure}

\subsection{Excited states via orthogonalisation}
\label{ssec:vqe_excited}

VQE in the form above targets only the ground state. To reach excited
levels, Gallimore and Liao constrain the trial state to be orthogonal to
the previously obtained eigenstates. Suppose the ground state
$\ket{\psi_0(\bm{\theta}_0^{\star})}$ has been determined; the first excited
state is then obtained by minimising
\begin{equation}
  \label{eq:vqe_excited}
  E_1 = \min_{\bm{\theta}\,:\,\bra{\psi_0}\psi(\bm{\theta})\rangle = 0}\;
        \bra{\psi(\bm{\theta})} H \ket{\psi(\bm{\theta})}.
\end{equation}
For the two-parameter UCC ansatz at $N=3$ this constrained optimisation is
one-dimensional: the orthogonal complement of $\ket{\psi_0}$ inside the
ansatz manifold is parameterised by a single angle $\gamma$, and the
optimisation reduces to a one-dimensional scan~\cite{gallimore2023}. The
orthogonal trial state takes the form
\begin{equation}
  \label{eq:orthogonal_state}
  \ket{\psi_1(\gamma)} =
    \cos\gamma\,\ket{\psi_0^{\perp,a}}
    + \sin\gamma\,\ket{\psi_0^{\perp,b}},
\end{equation}
where $\ket{\psi_0^{\perp,a}}, \ket{\psi_0^{\perp,b}}$ are two fixed states
orthogonal to $\ket{\psi_0}$ that span the rest of the real ansatz
manifold. Iterating the procedure, each new excited state is constrained
to be orthogonal to all previously obtained states.

\begin{remark}[Error propagation]
A useful property of the orthogonalisation method is that errors in the
ground-state estimate propagate in a bounded way: if the variational ground
state has an error $\delta_0 = E_0^{\mathrm{VQE}} - E_0$, the first excited
state estimate $E_1^{\mathrm{VQE}}$ satisfies
$|E_1^{\mathrm{VQE}} - E_1| \leq \delta_0$~\cite{gallimore2023}, so the
accuracy does not deteriorate as we move up the spectrum.
\end{remark}

\subsection{Extension to the spin-dependent N=4 case}
\label{ssec:gallimore_extension}

The Gallimore--Liao construction was originally applied to a
\emph{spin-independent} Cornell potential truncated to $N=3$, producing a
$3\times 3$ Hamiltonian, a JW Hamiltonian on three qubits, and a
two-parameter UCC ansatz. In this thesis we want to compare VQE with
VQITE (\cref{sec:woloshyn}) and TTITE (\cref{sec:yi}) on the same physical
system---the spin-dependent ${}^1S_0$, ${}^3S_1$, ${}^1P_1$ matrices of
\cref{ssec:matrices}. The Gallimore--Liao construction extends in two
respects to make this possible.

\begin{itemize}\setlength{\itemsep}{3pt}
  \item \textbf{Spin dependence.} The Pauli-string structure of
    \cref{eq:second_quantised} is unchanged when the Hamiltonian matrix
    carries an extra term from the contact interaction
    \cref{eq:hyperfine}: the matrix elements
    $H_{mn}^{(\ell,S)}$ themselves shift between channels, but the
    second-quantised expression and the resulting JW Pauli decomposition
    are formally identical. The spin enters only through the choice of
    coefficients.
  \item \textbf{Basis size.} Moving from $N=3$ to $N=4$ adds one more
    orbital, hence one more qubit and one more UCC parameter.
    \Cref{eq:jw_skip2} contributes the new long-range hops, yielding the
    full 17-term Pauli expansion of \cref{eq:jw_N4_full}. The UCC operator
    gains a third generator $a_3^{\dagger}a_0 - a_0^{\dagger}a_3$ and a
    third parameter, producing the three-parameter form of
    \cref{eq:ucc_state}.
\end{itemize}

The resulting four-qubit VQE algorithm is a faithful extension of
Gallimore and Liao's procedure; we are not introducing new ingredients,
only enlarging the basis and admitting the spin-dependent coefficients
already present in the underlying matrix. With this extension in place,
VQE can be applied to exactly the same $4\times 4$ matrices used by VQITE
and TTITE in the next two sections, enabling the apples-to-apples
comparison reported in the next chapter.


% ──────────────────────────────────────────────────────────────────
\section{Woloshyn --- Variational Imaginary-Time Evolution}
\label{sec:woloshyn}

Woloshyn~\cite{woloshyn2023} simulates the same spin-dependent charmonium
system studied in this thesis---the ${}^1S_0$, ${}^3S_1$, and ${}^1P_1$
channel matrices of \cref{ssec:matrices}---using the Variational
Imaginary-Time Evolution (VQITE) algorithm of McArdle et
al.~\cite{mcardle2019} and Yuan et al.~\cite{yuan2019}. Unlike VQE,
VQITE locates the ground state by tracking the (non-unitary)
imaginary-time Schr\"{o}dinger equation through a parameterised ansatz
manifold, deriving classical equations of motion from a variational
principle. Woloshyn uses the compact direct mapping for the qubit
encoding---two qubits for four basis states---together with a
three-parameter $R_y$-based ansatz, and also computes the M1 and E1
radiative transition amplitudes between channels. We reproduce all of
this material below.

\subsection{Imaginary-time evolution as a ground-state projector}
\label{ssec:ite}

The imaginary-time Schr\"{o}dinger equation is obtained from its real-time
counterpart by the Wick rotation $t \mapsto -i\tau$, with $\tau \geq 0$,
giving
\begin{equation}
  \label{eq:ite_eq}
  \frac{\mathrm{d}\ket{\psi(\tau)}}{\mathrm{d}\tau} = -H\ket{\psi(\tau)}.
\end{equation}
Its (unnormalised) solution is $e^{-H\tau}\ket{\psi(0)}$. Expanding the
initial state in the energy eigenbasis,
$\ket{\psi(0)} = \sum_k a_k\,\ket{E_k}$, gives
\begin{equation}
  \label{eq:ite_decomposition}
  e^{-H\tau}\ket{\psi(0)}
  = \sum_k a_k\,e^{-E_k\tau}\,\ket{E_k}.
\end{equation}
Factoring out the ground-state contribution and renormalising,
\begin{equation}
  \label{eq:ite_ground_state}
  \ket{\psi(\tau)}
  \;=\;
  \frac{a_0\,\ket{E_0} +
        \sum_{k>0} a_k\,e^{-(E_k - E_0)\tau}\,\ket{E_k}}
       {\sqrt{|a_0|^{2} + \sum_{k>0} |a_k|^{2} e^{-2(E_k - E_0)\tau}}}
  \;\xrightarrow[\tau\to\infty]{}\;
  \frac{a_0}{|a_0|}\,\ket{E_0},
\end{equation}
provided $a_0 = \bra{E_0}\psi(0)\rangle \neq 0$. The excited-state
amplitudes are exponentially suppressed at a rate set by the gap $E_k -
E_0$, so the algorithm reaches the ground state in a number of
imaginary-time steps proportional to $1/(E_1 - E_0)$.

To preserve the norm dynamically, it is convenient to subtract the
instantaneous energy,
$E_\tau \equiv \bra{\psi(\tau)} H \ket{\psi(\tau)}$, from $H$. The
\emph{normalised} imaginary-time equation
\begin{equation}
  \label{eq:ite_normalised}
  \frac{\mathrm{d}\ket{\psi(\tau)}}{\mathrm{d}\tau}
  = -\bigl(H - E_\tau\bigr)\ket{\psi(\tau)}
\end{equation}
preserves the norm of $\ket{\psi}$ and has the same long-$\tau$ limit as
\cref{eq:ite_eq}, but is more convenient to discretise variationally.

\begin{remark}[Non-unitarity]
The operator $e^{-H\tau}$ is Hermitian and positive definite, but it is
not unitary: $\bigl(e^{-H\tau}\bigr)^{\dagger}\,e^{-H\tau} = e^{-2H\tau}
\neq I$. Quantum circuits, by contrast, are unitary by construction.
Implementing \cref{eq:ite_normalised} on a quantum processor therefore
requires an indirect strategy; VQITE projects the exact non-unitary
dynamics onto a unitary parameterised submanifold, while \cref{sec:yi}
will encode it probabilistically using ancillas.
\end{remark}

\subsection{Direct (compact) encoding}
\label{ssec:direct_encoding}

Woloshyn adopts the most economical qubit encoding compatible with the
$N=4$ basis: each HO basis function is assigned to one computational-basis
state of a two-qubit register via the binary expansion of its index,
\begin{equation}
  \label{eq:direct_map}
  |n\rangle_{\mathrm{HO}}
  \;\longmapsto\;
  |\mathrm{bin}(n)\rangle_{\mathrm{qubit}},
  \qquad n = 0,1,2,3.
\end{equation}
Explicitly,
\begin{equation}
  |0\rangle \mapsto |00\rangle,\quad
  |1\rangle \mapsto |01\rangle,\quad
  |2\rangle \mapsto |10\rangle,\quad
  |3\rangle \mapsto |11\rangle.
\end{equation}
For $N=4$ this requires only $\lceil\log_2 N\rceil = 2$ qubits, and every
computational-basis state of the register corresponds to a physical HO
state---there are no spurious states and no symmetry constraint on the
ansatz.

The channel Hamiltonian $H^{(\ell,S)}$ is then literally a $4\times 4$
matrix acting on this register, and its Pauli decomposition follows
directly from \cref{eq:pauli_sum}. Applying that formula to the ${}^1S_0$
matrix of \cref{eq:H_1S0} yields, for example,
\begin{equation}
  \label{eq:direct_decomp_1S0}
  H^{({}^1S_0)} =
  c_{II}\,II + c_{IZ}\,IZ + c_{ZI}\,ZI + c_{ZZ}\,ZZ
  + c_{IX}\,IX + c_{XI}\,XI + c_{ZX}\,ZX + c_{XZ}\,XZ
  + c_{XX}\,XX + c_{YY}\,YY,
\end{equation}
i.e.\ ten Pauli terms, consistent with the absence of strings containing
an odd number of $Y$ factors for a real-symmetric operator. The numerical
coefficients $\{c_k\}$ are obtained by the trace formula of
\cref{eq:pauli_sum} and are real.

\begin{example}[Identity coefficient]
Setting $P_k = I\otimes I$ in \cref{eq:pauli_sum} gives the energy offset
\[
  c_{II} = \frac{1}{4}\Tr H^{({}^1S_0)}
         = \frac{1}{4}(0.9431 + 3.3365 + 5.4382 + 7.3451) \approx 4.266,
\]
which simply records the trace of the channel matrix and contributes only
a constant shift to the eigenvalues.
\end{example}

The direct encoding has two clear advantages over the JW encoding of
\cref{ssec:jw_encoding}: it uses the minimal number of qubits, and every
computational-basis state is physical, so the variational ansatz need not
respect any external constraint. Its main drawback is that the resulting
Pauli strings carry no obvious physical interpretation---they are simply
the coefficients of an abstract $4\times 4$ matrix.

\subsection{Three-parameter variational ansatz}
\label{ssec:woloshyn_ansatz}

For the two-qubit direct encoding, a state of the form
\begin{equation}
  \label{eq:real_4state}
  \ket{\psi} = c_0\ket{00} + c_1\ket{01} + c_2\ket{10} + c_3\ket{11},
  \qquad c_n\in\mathbb{R},\;\sum_n c_n^{2} = 1,
\end{equation}
has three independent real degrees of freedom (four amplitudes minus the
normalisation). Because the channel Hamiltonians of
\cref{eq:H_1S0,eq:H_3S1,eq:H_1P1} are real and symmetric, their
eigenstates can be chosen real, and a three-parameter circuit producing
the most general real state of the form \cref{eq:real_4state} is
therefore expressible enough. Woloshyn's circuit applies two single-qubit
$R_y$ rotations, a CNOT, and a third $R_y$:
\begin{equation}
  \label{eq:direct_ansatz_op}
  U(\bm{\theta}) =
  \bigl[I \otimes R_y(\theta_2)\bigr]\,\mathrm{CNOT}_{0\to1}\,
  \bigl[R_y(\theta_1) \otimes R_y(\theta_0)\bigr],
\end{equation}
with $\bm{\theta} = (\theta_0,\theta_1,\theta_2)$. The corresponding
circuit is shown in \cref{fig:woloshyn_ansatz}.

\begin{figure}[htbp]
\centering
\begin{quantikz}
  \lstick{$\ket{0}$} & \gate{R_y(\theta_0)} & \ctrl{1} & \qw                 & \qw \\
  \lstick{$\ket{0}$} & \gate{R_y(\theta_1)} & \targ{}  & \gate{R_y(\theta_2)} & \qw
\end{quantikz}
\caption{Three-parameter ansatz used with the direct encoding. The two
  initial $R_y$ rotations create an arbitrary product state, the CNOT
  entangles the qubits, and the final $R_y$ provides the third independent
  degree of freedom required to reach a general real four-amplitude
  state.}
\label{fig:woloshyn_ansatz}
\end{figure}

Acting on $\ket{00}$, this circuit yields amplitudes
\begin{equation}
  \label{eq:direct_ansatz_state}
  \ket{\psi(\bm{\theta})} =
  \begin{pmatrix}
    c_0(c_1 c_2 - s_1 s_2)\\
    s_0(s_1 c_2 - c_1 s_2)\\
    c_0(c_1 s_2 + s_1 c_2)\\
    s_0(s_1 s_2 + c_1 c_2)
  \end{pmatrix},
  \qquad
  c_n = \cos(\theta_n/2),\; s_n = \sin(\theta_n/2),
\end{equation}
which can be verified to cover the unit sphere in $\mathbb{R}^{4}$ as
$\bm{\theta}$ ranges over a domain of size $4\pi^{3}$, modulo redundancies.

\subsection{The McLachlan variational principle}
\label{ssec:mclachlan}

VQITE replaces the exact evolution \cref{eq:ite_normalised} by a
\emph{parameterised} trajectory $\ket{\psi(\bm{\theta}(\tau))}$ on the
ansatz manifold of \cref{ssec:woloshyn_ansatz}. The parameters
$\bm{\theta}(\tau)$ are evolved so that the parameterised state tracks
the exact one as closely as possible. We adopt the principle of
McLachlan~\cite{mclachlan1964}, which minimises at each instant the
distance between the actual time-derivative of the trial state and the
right-hand side of \cref{eq:ite_normalised}:
\begin{equation}
  \label{eq:mclachlan}
  \delta_{\dot{\bm{\theta}}}\,
  \bigl\| \tfrac{\partial}{\partial\tau}\ket{\psi(\bm{\theta}(\tau))}
        + (H - E_\tau)\ket{\psi(\bm{\theta}(\tau))} \bigr\|^{2} = 0.
\end{equation}
Here the variation is taken with respect to the time derivative
$\dot{\bm{\theta}} \equiv \mathrm{d}\bm{\theta}/\mathrm{d}\tau$, treated
as an independent real vector. The variation reduces, in five short
steps, to the linear system
\begin{equation}
  \label{eq:mclachlan_linsys}
  \sum_j A_{ij}\,\dot{\theta}_j = C_i,
  \qquad i = 1,\ldots,p,
\end{equation}
with matrix and vector elements
\begin{align}
  \label{eq:mclachlan_A}
  A_{ij} &= \Re\bigl[
    \braket{\partial_i\psi}{\partial_j\psi}
    - \braket{\partial_i\psi}{\psi}\braket{\psi}{\partial_j\psi}
  \bigr],\\[2pt]
  \label{eq:mclachlan_C}
  C_i &= -\Re\bigl[\bra{\partial_i\psi} H \ket{\psi}\bigr]
       + \Re\bigl[\braket{\partial_i\psi}{\psi}\bigr]\,E_\tau.
\end{align}
We work through the derivation in detail since the same building blocks
will reappear when we discuss the metric tensor and the parameter-shift
rule below.

\paragraph{Step 1 --- expand the time derivative.}
The ansatz state depends on $\tau$ only through the parameters, so by
the chain rule
\begin{equation}
  \label{eq:dtau_psi}
  \partial_\tau\ket{\psi(\bm{\theta}(\tau))}
  = \sum_i \dot{\theta}_i\,\ket{\partial_i\psi},
  \qquad
  \ket{\partial_i\psi} \equiv \frac{\partial\ket{\psi(\bm{\theta})}}
                                   {\partial\theta_i}.
\end{equation}

\paragraph{Step 2 --- write the residual vector.}
Denote $K \equiv H - E_\tau$ for brevity. The vector inside the norm in
\cref{eq:mclachlan} is
\begin{equation}
  \label{eq:residual}
  \ket{R} \;=\; \sum_i \dot{\theta}_i\,\ket{\partial_i\psi}
                + K\,\ket{\psi}.
\end{equation}

\paragraph{Step 3 --- compute its squared norm.}
Expanding $\braket{R}{R}$,
\begin{equation*}
  \|R\|^{2} =
    \sum_{i,j}\dot{\theta}_i\dot{\theta}_j\,
       \braket{\partial_i\psi}{\partial_j\psi}
    + \sum_i \dot{\theta}_i\,\bra{\partial_i\psi}K\ket{\psi}
    + \sum_j \dot{\theta}_j\,\bra{\psi}K\ket{\partial_j\psi}
    + \bra{\psi}K^{2}\ket{\psi}.
\end{equation*}
The two middle sums are complex conjugates of each other (since $K$ is
Hermitian and $\dot{\theta}_i\in\mathbb{R}$); they combine into a single
real part. In the first sum the antisymmetric imaginary part of
$\braket{\partial_i\psi}{\partial_j\psi}$ vanishes against the symmetric
$\dot{\theta}_i\dot{\theta}_j$, leaving only the real part. Hence
\begin{equation}
  \label{eq:residual_norm}
  \|R\|^{2} =
    \sum_{i,j}\dot{\theta}_i\dot{\theta}_j\,
      \Re\braket{\partial_i\psi}{\partial_j\psi}
    + 2\sum_i \dot{\theta}_i\,
      \Re\bra{\partial_i\psi}K\ket{\psi}
    + \|K\ket{\psi}\|^{2}.
\end{equation}

\paragraph{Step 4 --- vary with respect to $\dot{\theta}_k$.}
The final term in \cref{eq:residual_norm} is independent of
$\dot{\bm{\theta}}$. Differentiating the remainder with respect to
$\dot{\theta}_k$ and setting the derivative to zero,
\begin{equation}
  \label{eq:mclachlan_variation}
  \frac{\partial\|R\|^{2}}{\partial\dot{\theta}_k}
  = 2\sum_j \Re\braket{\partial_k\psi}{\partial_j\psi}\,\dot{\theta}_j
    + 2\,\Re\bra{\partial_k\psi}K\ket{\psi}
  \;=\; 0,
\end{equation}
which is already linear in $\dot{\bm{\theta}}$, matching the structure
of \cref{eq:mclachlan_linsys} with the \emph{bare} entries
\begin{align}
  A^{(0)}_{kj} &= \Re\braket{\partial_k\psi}{\partial_j\psi},
  \label{eq:A_bare}\\
  C_k &= -\Re\bra{\partial_k\psi}K\ket{\psi}
       = -\Re\bra{\partial_k\psi}H\ket{\psi}
         + E_\tau\,\Re\braket{\partial_k\psi}{\psi}.
  \label{eq:C_from_K}
\end{align}
The right-hand side of \cref{eq:C_from_K} is precisely
\cref{eq:mclachlan_C}, so the vector $C_i$ is recovered at this step
already.

\paragraph{Step 5 --- project to the Fubini--Study form.}
The bare metric \cref{eq:A_bare} misses one subtlety: the parameterised
state has a redundant global-phase degree of freedom---$\ket{\psi(\bm{\theta})}$
and $e^{i\phi(\bm{\theta})}\ket{\psi(\bm{\theta})}$ represent the same
physical state, but their tangent vectors differ by an arbitrary
multiple of $\ket{\psi}$. To remove this gauge redundancy we project the
tangent vectors orthogonal to $\ket{\psi}$,
\begin{equation}
  \label{eq:proj_tangent}
  \ket{\partial_i\psi}_{\perp}
  \;\equiv\;
  \bigl(\,I - \ket{\psi}\bra{\psi}\,\bigr)\,\ket{\partial_i\psi}
  = \ket{\partial_i\psi} - \ket{\psi}\,\braket{\psi}{\partial_i\psi},
\end{equation}
and use $\ket{\partial_i\psi}_{\perp}$ in place of
$\ket{\partial_i\psi}$ throughout. A direct computation using
$\braket{\psi}{\psi}=1$ gives
\begin{equation*}
  \braket{\partial_i\psi|_{\perp}}{\partial_j\psi|_{\perp}}
  = \braket{\partial_i\psi}{\partial_j\psi}
   - \braket{\partial_i\psi}{\psi}\,\braket{\psi}{\partial_j\psi},
\end{equation*}
whose real part is exactly \cref{eq:mclachlan_A}. The same projection
on $\ket{\partial_k\psi}$ in $C_k$ leaves it unchanged, because
$\bra{\psi}K\ket{\psi} = E_\tau - E_\tau = 0$ already; only $A$ acquires
the correction.
For ansatz circuits that preserve the global phase of $\ket{\psi}$ (which
is the case for the construction of \cref{ssec:woloshyn_ansatz}) the
second term in $C_i$ vanishes and $C_i$ reduces to a real-valued energy
gradient,
\begin{equation}
  \label{eq:mclachlan_C_simplified}
  C_i = -\frac{1}{2}\,\frac{\partial E(\bm{\theta})}{\partial\theta_i}.
\end{equation}
The matrix $A$ is precisely the (real) Fubini--Study quantum metric on the
ansatz manifold; it is real, symmetric, and positive semidefinite, and it
plays the same role for variational quantum evolution as the Fisher
information matrix in classical statistics~\cite{stokes2020}.

\subsection{The quantum metric tensor in practice}
\label{ssec:metric_tensor}

The geometric interpretation of $A_{ij}$ is illuminating, but more useful
for implementation is its expression as a curvature of the
\emph{fidelity surface}
\begin{equation}
  \label{eq:fidelity_func}
  f(\bm{\theta}, \bm{\theta}')
  \equiv |\braket{\psi(\bm{\theta})}{\psi(\bm{\theta}')}|^{2}.
\end{equation}
Expanding around $\bm{\theta}' = \bm{\theta}$ to second order yields
\begin{equation}
  \label{eq:fidelity_curvature}
  f(\bm{\theta}, \bm{\theta} + \bm{\delta})
  = 1 - \sum_{ij} A_{ij}\,\delta_i\delta_j + O(\delta^{3}),
\end{equation}
so that the metric coincides with the leading curvature of the fidelity
landscape. This identification is the basis of the standard
finite-difference estimator: from \cref{eq:fidelity_curvature},
\begin{equation}
  \label{eq:metric_fd}
  A_{ij} \approx -\frac{1}{4\varepsilon^{2}}\,
  \bigl[\,
    f(\bm{\theta} + \varepsilon e_i + \varepsilon e_j)
    - f(\bm{\theta} + \varepsilon e_i - \varepsilon e_j)
    - f(\bm{\theta} - \varepsilon e_i + \varepsilon e_j)
    + f(\bm{\theta} - \varepsilon e_i - \varepsilon e_j)
  \bigr],
\end{equation}
with $e_i$ the unit vector in the $i$-direction. Each of the four
fidelities is the absolute square of an overlap between two ansatz
states, which can be estimated on a quantum processor either by the swap
test or by the inverse-ansatz protocol described later in
\cref{ssec:transitions}. For the three-parameter ansatz used here, the
symmetric metric requires $p(p+1)/2 = 6$ independent entries, each
costing four overlap measurements at a small step
$\varepsilon \sim 10^{-3}$--$10^{-4}$.

\subsection{Energy gradient and the parameter-shift rule}
\label{ssec:parameter_shift}

The vector $C_i$ in \cref{eq:mclachlan_C_simplified} is a derivative of
an expectation value with respect to a circuit parameter. For ansatze
composed of single-qubit rotations of the form
$R_\mu(\theta) = e^{-i\theta\,\mu/2}$ with $\mu \in \{X,Y,Z\}$, such
derivatives admit a remarkable closed-form expression---the
\emph{parameter-shift rule}~\cite{mitarai2018,schuld2019}:
\begin{equation}
  \label{eq:param_shift}
  \frac{\partial \langle H\rangle_{\bm{\theta}}}{\partial\theta_i}
  =
  \frac{1}{2}\Bigl[
    \langle H\rangle_{\bm{\theta} + (\pi/2)\,e_i}
  - \langle H\rangle_{\bm{\theta} - (\pi/2)\,e_i}
  \Bigr].
\end{equation}
The remarkable feature of \cref{eq:param_shift} is that it is
\emph{exact}---unlike a finite-difference derivative, the formula returns
the true partial derivative regardless of the step size. The proof, which
we sketch here, illustrates the special role played by Pauli rotations.

\begin{remark}[Sketch of the parameter-shift derivation]
Consider a circuit where the parameter $\theta_i$ appears only inside a
rotation $R_\mu(\theta_i) = e^{-i\theta_i \mu/2}$. Since $\mu^{2}=I$, we
have $R_\mu(\theta_i) = \cos(\theta_i/2)\,I - i\sin(\theta_i/2)\,\mu$, so
the dependence of $R_\mu$---and hence of
$\ket{\psi(\bm{\theta})}$---on $\theta_i$ is restricted to the linear
span of $\{1,\cos\theta_i,\sin\theta_i\}$. Therefore the expectation
value $\langle H\rangle_{\bm{\theta}}$ is a real-valued trigonometric
polynomial of period $2\pi$ in $\theta_i$, of the form
\(
  a\cos\theta_i + b\sin\theta_i + c.
\)
Setting the right-hand side of \cref{eq:param_shift} equal to
$b\cos\theta_i - a\sin\theta_i$, which is exactly the derivative of
$a\cos\theta_i + b\sin\theta_i + c$, completes the proof.
\end{remark}

The practical implication is that, for the $R_y$-based ansatz of
\cref{eq:direct_ansatz_op}, all partial derivatives required by
VQITE---i.e.\ each component of $C$---can be evaluated on the quantum
processor at a cost of two additional energy estimations per parameter.
For the three-parameter ansatz used here, the cost is just six extra
circuit evaluations per imaginary-time step.

\subsection{Time stepping and regularisation}
\label{ssec:vqite_stepping}

Combining \cref{eq:mclachlan_linsys,eq:mclachlan_A,eq:mclachlan_C_simplified}
gives the equation of motion for $\bm{\theta}$:
\begin{equation}
  \label{eq:vqite_eom}
  \dot{\bm{\theta}} = A^{-1}\,C.
\end{equation}
Discretising with the forward Euler scheme yields a simple update rule
for the parameter vector after a small imaginary-time step $\Delta\tau$,
\begin{equation}
  \label{eq:vqite_update}
  \bm{\theta}(\tau + \Delta\tau)
  = \bm{\theta}(\tau)
  + \Delta\tau\,A^{-1}(\bm{\theta})\,C(\bm{\theta}).
\end{equation}
Two practical issues arise. First, when two parameters become
approximately degenerate (e.g., when the ansatz fails to be locally
injective), the metric $A$ becomes ill-conditioned. We add a small
Tikhonov regulariser to restore numerical stability,
\begin{equation}
  \label{eq:vqite_tikhonov}
  (A + \lambda I)\,\dot{\bm{\theta}} = C,
  \qquad \lambda \sim 10^{-4}.
\end{equation}
Second, the explicit Euler scheme is conditionally stable and accuracy is
limited by the curvature of the energy landscape. Step sizes around
$\Delta\tau \simeq 0.02$ in natural units of $1/\|H\|$ are sufficient to
reach a stable ground state in $50$--$100$ steps for the channel
Hamiltonians of \cref{eq:H_1S0,eq:H_3S1,eq:H_1P1}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.2cm and 2.2cm,
    box/.style={
      draw, rounded corners=4pt, minimum width=3.5cm,
      minimum height=0.9cm, align=center, font=\small
    },
    arrow/.style={-{Stealth[length=6pt]}, thick},
  ]
  \node[box, fill=blue!8] (prep) {Prepare $\ket{\psi(\bm{\theta})}$};
  \node[box, fill=blue!8, right=of prep] (meas)
    {Estimate $A_{ij},\,C_i$\\
     {\footnotesize (overlaps + parameter shifts)}};
  \node[box, fill=orange!10, below=of meas] (solve)
    {Solve $(A+\lambda I)\dot{\bm{\theta}} = C$};
  \node[box, fill=orange!10, below=of prep] (upd)
    {Update $\bm{\theta} \mathrel{+}= \Delta\tau\,\dot{\bm{\theta}}$};
  \draw[arrow] (prep) -- (meas);
  \draw[arrow] (meas) -- (solve);
  \draw[arrow] (solve) -- (upd);
  \draw[arrow] (upd) -- (prep);
  \node[font=\footnotesize\itshape, above=0.1cm of prep, anchor=south west,
        xshift=-0.15cm] {Quantum};
  \node[font=\footnotesize\itshape, below=0.1cm of upd, anchor=north west,
        xshift=-0.15cm] {Classical};
\end{tikzpicture}
\caption{The VQITE hybrid loop. At each step the quantum processor
  estimates the metric tensor $A$ via \cref{eq:metric_fd} and the energy
  gradient components of $C$ via the parameter-shift rule
  \cref{eq:param_shift}; the classical computer solves the linear system
  \cref{eq:vqite_tikhonov} and advances the parameters in imaginary time.}
\label{fig:vqite_loop}
\end{figure}

\subsection{Excited states by penalisation}
\label{ssec:vqite_excited}

VQITE projects onto the lowest-energy state with non-vanishing overlap
with the initial state. To target higher levels we modify the effective
Hamiltonian by adding penalty terms that suppress overlap with previously
obtained states. Concretely, if $\{\ket{\psi_k^{\star}}\}_{k=0}^{n-1}$
are already-computed eigenstates, we replace $H$ by
\begin{equation}
  \label{eq:penalised_H}
  H_{\mathrm{eff}}
  = H + \alpha\sum_{k=0}^{n-1} \ket{\psi_k^{\star}}\bra{\psi_k^{\star}},
\end{equation}
where $\alpha$ is a positive penalty large enough to shift the energies
of the already-found states above the next eigenvalue. Running VQITE
with $H_{\mathrm{eff}}$ then converges to the next-higher eigenstate.
The energy gradient generalises to
\begin{equation}
  \label{eq:penalised_grad}
  \frac{\partial \langle H_{\mathrm{eff}}\rangle}{\partial\theta_i}
  = \frac{\partial \langle H\rangle}{\partial\theta_i}
  + 2\alpha\sum_{k}\Re\bigl[\braket{\psi_k^{\star}}{\psi}\,
                            \bra{\partial_i\psi}\psi_k^{\star}\rangle\bigr],
\end{equation}
where the overlap correction is again accessible from short fidelity
circuits. In practice $\alpha$ is chosen of the same order of magnitude
as the typical gap between adjacent eigenvalues; we use $\alpha = 10$ in
inverse-femtometre units throughout this thesis.

The combination of VQITE for the ground state with iterated penalisation
provides access to a sequence of eigenpairs without ever requiring the
diagonalisation of an exponentially large matrix on a classical computer.

\subsection{Transition amplitudes}
\label{ssec:transitions}

Woloshyn's analysis goes beyond eigenvalues: it also computes radiative
transition amplitudes between channels, which are the physical
observables ultimately compared with experiment. We reproduce the two
relevant cases below.

\paragraph{Physical motivation.}
In the multipole expansion of the electromagnetic interaction of a
quark--antiquark system, the leading single-photon transitions are
classified by their selection rules~\cite{brambilla2011}:
\begin{itemize}\setlength{\itemsep}{2pt}
  \item \emph{M1 transitions} flip the total spin without changing the
    orbital angular momentum ($\Delta S = 1$, $\Delta L = 0$). In the
    non-relativistic limit the amplitude reduces to a wavefunction
    overlap between the two spin partners, scaled by a numerical
    prefactor.
  \item \emph{E1 transitions} change the orbital angular momentum by one
    unit without altering the spin ($\Delta S = 0$, $\Delta L = 1$). The
    amplitude is proportional to the radial matrix element of the
    position operator, $\bra{\psi_f}r\ket{\psi_i}$.
\end{itemize}
The two paradigmatic examples for charmonium are
$J/\psi \to \eta_c\,\gamma$ (M1) and $h_c \to \eta_c\,\gamma$ (E1).
Computing their amplitudes therefore amounts to evaluating, respectively,
the overlap $\braket{\psi_f^{({}^1S_0)}}{\psi_i^{({}^3S_1)}}$ and the
matrix element $\bra{\psi_f^{({}^1S_0)}}r\ket{\psi_i^{({}^1P_1)}}$, where
the states are the eigenvectors produced by VQITE (or by any of the
other two algorithms).

\paragraph{M1 amplitudes: state overlaps.}
In the non-relativistic limit, the M1 transition amplitude between two
states differing only in spin reduces to an overlap of their spatial
wavefunctions,
\begin{equation}
  \label{eq:m1_amplitude}
  |\mathcal{M}_{\mathrm{M1}}|^{2} \;\propto\;
  \bigl|\braket{\psi_f^{\mathrm{spatial}}}{\psi_i^{\mathrm{spatial}}}\bigr|^{2}.
\end{equation}
The two states are eigenvectors of different channel Hamiltonians
(${}^1S_0$ and ${}^3S_1$) but live in the same HO basis, so their
overlap is well-defined as a four-dimensional inner product of real
coefficient vectors. Two complementary quantum-circuit techniques are
used in Woloshyn's analysis.

\textit{Inverse-ansatz protocol.}
If the two ansatz unitaries $U_i = U(\bm{\theta}_i^{\star})$ and
$U_f = U(\bm{\theta}_f^{\star})$ have been determined---e.g., as the output
of two VQITE runs---then the overlap can be read off from the all-zero
outcome of the circuit $U_f^{\dagger} U_i\ket{0}^{\otimes n}$. Indeed,
\begin{equation}
  \label{eq:inverse_ansatz}
  \mathrm{Prob}\bigl(\text{all zeros}\bigr) =
    \bigl|\bra{0}^{\otimes n}\,U_f^{\dagger} U_i\,\ket{0}^{\otimes n}\bigr|^{2}
  = \bigl|\braket{\psi_f}{\psi_i}\bigr|^{2}.
\end{equation}
This protocol requires no additional qubits, only an inverse-ansatz
sub-circuit, and is the most economical method available when both
ansatz unitaries can be efficiently implemented.

\textit{Swap test.}
A more general protocol, applicable when $\ket{\psi_i}$ and
$\ket{\psi_f}$ are prepared independently on two copies of the work
register, is the swap test~\cite{nielsen2010}. Adding one ancilla
initialised in $\ket{0}$, the circuit shown in \cref{fig:swap_test}
yields an ancilla-zero probability
\(
  \mathrm{Prob}(0_{\mathrm{anc}})
   = \tfrac{1}{2}\bigl(1 + |\braket{\psi_f}{\psi_i}|^{2}\bigr),
\)
from which the overlap squared is recovered as
$2\,\mathrm{Prob}(0_{\mathrm{anc}}) - 1$. The swap test is more demanding
in qubits ($2n + 1$ total) and gates ($n$ controlled-SWAPs), but does
not require explicit knowledge of either ansatz unitary, only the
ability to prepare both states on demand.

\begin{figure}[htbp]
\centering
\begin{quantikz}[column sep=10pt]
  \lstick{$\ket{0}_{\mathrm{anc}}$} & \gate{H} & \ctrl{1}  & \gate{H} & \meter{} \\
  \lstick{$\ket{\psi_i}$}           & \qw      & \swap{1}  & \qw      & \qw \\
  \lstick{$\ket{\psi_f}$}           & \qw      & \targX{}  & \qw      & \qw
\end{quantikz}
\caption{Swap test for the absolute overlap between two ansatz states
  $\ket{\psi_i}$ and $\ket{\psi_f}$. The Hadamard--CSWAP--Hadamard
  sandwich interferes the swapped and unswapped configurations of the
  work registers; the ancilla-zero probability is
  $(1 + |\braket{\psi_f}{\psi_i}|^{2})/2$. The CSWAP gate acts as a
  controlled exchange of the two work registers.}
\label{fig:swap_test}
\end{figure}

\paragraph{E1 amplitudes: matrix elements of the radial operator.}
For E1 transitions the relevant matrix element is the radial dipole
amplitude $\bra{\psi_f}r\ket{\psi_i}$, where $\ket{\psi_i}$ and
$\ket{\psi_f}$ are eigenstates of the initial- and final-state
channels. In the HO basis the radial operator is an $N\times N$ matrix
$R_{mk} = \bra{\phi_{m\,\ell_f}} r \ket{\phi_{k\,\ell_i}}$ which, like
the Hamiltonian, can be evaluated in closed form. Using the same
techniques that produced \cref{eq:linear_matrix}, one obtains, for the
$P\to S$ case ($\ell_i = 1, \ell_f = 0$) and $N = 4$, the explicit
matrix~\cite{woloshyn2023}
\begin{equation}
  \label{eq:r_matrix}
  R \;=\;
  \begin{pmatrix}
    0.5775 & -0.4715 & 0      & 0      \\
    0      &  0.8821 & -0.6668 & 0     \\
    0      &  0      &  0.8821 & -0.8167\\
    0      &  0      &  0      &  1.0002
  \end{pmatrix}
  \quad \text{(units of fm)}.
\end{equation}
The amplitude is then
\begin{equation}
  \label{eq:e1_amplitude}
  \bra{\psi_f}r\ket{\psi_i} = \mathbf{c}_f^{\!\top} R\,\mathbf{c}_i,
\end{equation}
where $\mathbf{c}_i$ and $\mathbf{c}_f$ are the real coefficient vectors
of $\ket{\psi_i}$ and $\ket{\psi_f}$ in the HO basis.

\begin{remark}[Errata]
Woloshyn's Eq.~(13) reports the $(2,2)$ element of $R$ as $5.4382$, which
is in fact the numerical value of an entry in the ${}^1S_0$ Hamiltonian
matrix \cref{eq:H_1S0}; the correct value, consistent with the
monotonically growing radial-element pattern, is $0.8821$, as used in
\cref{eq:r_matrix}. This is one of two copy-paste errata noted in this
chapter (the other being the ${}^1P_1$ diagonal entry of
\cref{ssec:matrices}).
\end{remark}

To extract $\bra{\psi_f}r\ket{\psi_i}$ on a quantum device, we first
Pauli-decompose the operator $R$ on the work register (analogously to
\cref{ssec:pauli_decomp}). The matrix element then reduces to a sum of
terms of the form $\bra{\psi_f}P\ket{\psi_i}$ for Pauli strings $P$.
Each such term can be estimated by the \emph{Hadamard
test}~\cite{nielsen2010}: with one ancilla qubit prepared in $\ket{0}$,
the circuit shown in \cref{fig:hadamard_test} yields an ancilla-zero
probability
\begin{equation}
  \label{eq:hadamard_test}
  \mathrm{Prob}(0_{\mathrm{anc}})
  = \frac{1 + \Re\,\bra{\psi_f}P\ket{\psi_i}}{2},
\end{equation}
from which the real part of the matrix element is obtained as
$2\,\mathrm{Prob}(0_{\mathrm{anc}}) - 1$. Replacing the leading
Hadamard on the ancilla by a $S^{\dagger} H$ rotation yields, by the same
algebra, the imaginary part of the matrix element. For real-symmetric
problems such as the channel Hamiltonians considered here, only the real
part is needed.

\begin{figure}[htbp]
\centering
\begin{quantikz}[column sep=10pt]
  \lstick{$\ket{0}_{\mathrm{anc}}$} & \gate{H} & \ctrl{1}     & \gate{H} & \meter{} \\
  \lstick{$\ket{\psi_i}$}           & \gate{U_{f\leftarrow i}^{\dagger}\,P} & \qw & \qw & \qw
\end{quantikz}
\caption{Hadamard test for the matrix element $\bra{\psi_f}P\ket{\psi_i}$.
  Conditioned on the ancilla, the work register is rotated by an operator
  whose expectation value coincides with the desired matrix element. The
  ancilla-zero probability is $(1 + \Re\bra{\psi_f}P\ket{\psi_i})/2$.
  Replacing the first Hadamard by $S^{\dagger}H$ gives the imaginary
  part.}
\label{fig:hadamard_test}
\end{figure}

The E1 transition amplitude is then a weighted sum of Hadamard tests,
one per Pauli string in the decomposition of $R$:
\begin{equation}
  \label{eq:e1_decomposed}
  \bra{\psi_f}r\ket{\psi_i}
  = \sum_k r_k\,\bra{\psi_f}P_k\ket{\psi_i},
\end{equation}
where $r_k$ are the Pauli coefficients of $R$ obtained from
\cref{eq:pauli_sum}. With this construction, the same quantum-algorithmic
framework that yields energies also yields the matrix elements needed
for radiative transitions, allowing a direct comparison with
experimental decay widths and lifetimes once eigenstates have been
obtained.


% ──────────────────────────────────────────────────────────────────
\section{Yi et al.\ --- Trotter--Taylor Imaginary-Time Evolution}
\label{sec:yi}

Yi et al.~\cite{yi2025} introduce a probabilistic imaginary-time evolution
algorithm, named Trotter--Taylor Imaginary-Time Evolution (TTITE), based
on combining a Trotter decomposition of the imaginary-time propagator
with a truncated Taylor expansion implemented through a linear
combination of unitaries (LCU). They demonstrate the algorithm on the
molecular hydrogen ($\mathrm{H}_{2}$) Hamiltonian and on one-dimensional
Heisenberg spin chains. The construction itself depends only on the
Pauli decomposition of the target Hamiltonian and is therefore generic;
in this thesis we apply it to the channel Hamiltonians of
\cref{ssec:matrices} in the same direct encoding used by Woloshyn
(\cref{ssec:direct_encoding}), providing a third independent route to
the charmonium spectrum.

\subsection{Strategy: implement the propagator directly}
\label{ssec:ttite_strategy}

In contrast to VQE and VQITE, TTITE does not fit any ansatz. The starting
point is again the eigenstate-projection property of imaginary time
\cref{eq:ite_ground_state}. We seek a unitary protocol that, given the
initial state $\ket{\psi(0)}$ on the work register, returns
\(
  \ket{\psi(\tau)} \propto e^{-H\tau}\ket{\psi(0)}
\)
to controllable accuracy. Two obstacles must be overcome.

The first obstacle is the non-unitarity of the propagator
$e^{-H\tau}$. We resolve it by writing each elementary factor of a
Trotter-decomposed propagator as a linear combination of two unitaries;
such ``LCU'' operators admit a probabilistic implementation via an
ancilla.

The second obstacle is that the Hamiltonian is a \emph{sum} of Pauli
strings, and exponentials of a sum cannot generally be separated. We
resolve it through the Trotter formula, which decomposes
$e^{-H\Delta\tau}$ into a product of exponentials of individual Pauli
strings, each of which is straightforward to handle.

We now make each of these steps precise.

\subsection{Trotter decomposition}
\label{ssec:trotter}

Let $H = \sum_{i=1}^{m} c_i\, h_i$ be the Pauli decomposition of one of
the channel Hamiltonians (obtained via \cref{eq:pauli_sum}, with the
direct encoding), with $c_i\in\mathbb{R}$ and each $h_i$ a Pauli string
satisfying $h_i^{2} = I$. The first-order \emph{Trotter
formula}~\cite{trotter1959,suzuki1976} approximates the imaginary-time
propagator for a small step $\Delta\tau$ as
\begin{equation}
  \label{eq:trotter}
  e^{-H\,\Delta\tau}
  = \prod_{i=1}^{m} e^{-c_i\,h_i\,\Delta\tau}
  + O\!\bigl((\Delta\tau)^{2}\bigr),
\end{equation}
with leading error proportional to the commutators $[h_i, h_j]$. To
evolve over a total imaginary time $\tau$ we apply $L = \tau/\Delta\tau$
such Trotter steps in succession, accumulating an overall error of order
$\tau\,\Delta\tau$. Higher-order symmetric (Suzuki) schemes reduce this
error to $O(\Delta\tau^{2k+1})$ for stage-$2k$ formulas, at the cost of
more elementary exponentials per step~\cite{suzuki1976}; for the small
Hamiltonians of this work the first-order formula \cref{eq:trotter} is
already accurate enough.

\subsection{Taylor expansion of a single Trotter step}
\label{ssec:taylor_step}

Each factor $e^{-c\,h\,\Delta\tau}$ in \cref{eq:trotter} is the
exponential of an Hermitian operator whose square is the identity,
$h^{2} = I$. This property gives a remarkably simple closed form.
Expanding the exponential as a power series and separating even and odd
powers,
\begin{align}
  \label{eq:taylor_series}
  e^{-c\,h\,\Delta\tau}
  &= \sum_{j=0}^{\infty} \frac{(-c\,\Delta\tau)^{j}}{j!}\,h^{j}\\
  &= \biggl(\sum_{j\text{ even}} \frac{(-c\,\Delta\tau)^{j}}{j!}\biggr)\,I
   + \biggl(\sum_{j\text{ odd}}  \frac{(-c\,\Delta\tau)^{j}}{j!}\biggr)\,h
   = \cosh(c\,\Delta\tau)\,I - \sinh(c\,\Delta\tau)\,h,
  \label{eq:closed_form}
\end{align}
where we used $h^{2k} = I$ and $h^{2k+1} = h$. The single Trotter step
is therefore exactly a linear combination of two unitaries: the identity
and $h$ itself. Truncating the series at finite order $R$,
\begin{equation}
  \label{eq:taylor_truncation}
  T^{(R)}(c\,\Delta\tau)
  \;\equiv\; \alpha\,I + \beta\,h,
  \qquad
  \alpha = \sum_{\substack{j=0\\j\text{ even}}}^{R}\!\frac{(-c\,\Delta\tau)^{j}}{j!},
  \qquad
  \beta  = \sum_{\substack{j=1\\j\text{ odd}}}^{R}\!\frac{(-c\,\Delta\tau)^{j}}{j!},
\end{equation}
incurs a Taylor error of order $(\Delta\tau)^{R+1}$, which for
$R = 4$ or higher is negligible compared to the Trotter error
$O(\Delta\tau^{2})$ in \cref{eq:trotter}. The truncation is convenient
because it makes $\alpha$ and $\beta$ polynomials in $\Delta\tau$,
slightly simpler than $\cosh$ and $\sinh$.

\begin{remark}
The closed form \cref{eq:closed_form} shows that the Taylor truncation
is in fact \emph{not necessary} for correctness; one could simply use
$\alpha = \cosh(c\,\Delta\tau)$ and $\beta = -\sinh(c\,\Delta\tau)$. The
finite-order form \cref{eq:taylor_truncation} matches the construction
of Yi et al.~\cite{yi2025} and makes the analysis of the LCU success
probability slightly more uniform across different Hamiltonians, but the
algorithm itself is unchanged.
\end{remark}

\subsection{Linear-combination-of-unitaries (LCU) implementation}
\label{ssec:lcu}

The operator $\alpha I + \beta h$ in \cref{eq:taylor_truncation} is a
non-unitary linear combination of two unitaries. Operators of this form
admit a celebrated probabilistic implementation, the LCU
construction~\cite{childs2012}, which uses a single ancillary qubit. We
derive it explicitly here.

Augment the work register with one ancilla qubit, initially in $\ket{0}$.
The combined state is $\ket{0}\otimes\ket{\psi}$. The LCU circuit
applies the following three operations.

\paragraph{(i) Ancilla preparation.}
Apply $R_y(\vartheta)$ to the ancilla with
\begin{equation}
  \label{eq:lcu_angle}
  \vartheta = 2\atantwo(\beta,\alpha),
\end{equation}
so that
\(
  R_y(\vartheta)\ket{0}
  = \cos(\vartheta/2)\ket{0} + \sin(\vartheta/2)\ket{1}
  = (\alpha\ket{0} + \beta\ket{1})/\sqrt{\alpha^{2}+\beta^{2}}.
\)
The combined state is then
\begin{equation}
  \label{eq:lcu_state1}
  \ket{\phi_1} = \frac{1}{\sqrt{\alpha^{2}+\beta^{2}}}
                 \bigl(\alpha\ket{0} + \beta\ket{1}\bigr)\otimes\ket{\psi}.
\end{equation}

\paragraph{(ii) Controlled application of $h$.}
Apply the controlled unitary
\(
  CU = \ket{0}\bra{0}\otimes I + \ket{1}\bra{1}\otimes h,
\)
which acts as the identity if the ancilla is $\ket{0}$ and as $h$ if it
is $\ket{1}$. Since $h$ is a Pauli string of length $n$, the controlled
gate decomposes into $n$ elementary controlled-Pauli gates acting on the
work qubits. After this step,
\begin{equation}
  \label{eq:lcu_state2}
  \ket{\phi_2}
  = \frac{1}{\sqrt{\alpha^{2}+\beta^{2}}}
    \bigl(\alpha\ket{0}\otimes\ket{\psi}
        + \beta\ket{1}\otimes h\ket{\psi}\bigr).
\end{equation}

\paragraph{(iii) Hadamard and measurement on the ancilla.}
Apply a Hadamard gate to the ancilla and then measure it in the
computational basis. Using $H\ket{0} = (\ket{0}+\ket{1})/\sqrt{2}$,
$H\ket{1} = (\ket{0}-\ket{1})/\sqrt{2}$, the state just before
measurement is
\begin{equation}
  \label{eq:lcu_state3}
  \ket{\phi_3}
  = \frac{1}{\sqrt{2(\alpha^{2}+\beta^{2})}}
    \Bigl(\ket{0}\otimes(\alpha I + \beta h)\ket{\psi}
        + \ket{1}\otimes(\alpha I - \beta h)\ket{\psi}\Bigr).
\end{equation}
A measurement outcome of $0$ on the ancilla projects the work register
onto the (unnormalised) state $(\alpha I + \beta h)\ket{\psi}$. After
renormalisation this is precisely the desired truncated Trotter step.

The probability of measuring $\ket{0}$ on the ancilla,
\begin{equation}
  \label{eq:lcu_success}
  P_{\mathrm{s}} =
  \frac{\bigl\|(\alpha I + \beta h)\ket{\psi}\bigr\|^{2}}
       {2(\alpha^{2} + \beta^{2})},
\end{equation}
sets the cost of the protocol: if the measurement returns $\ket{1}$ the
work state has been corrupted and the segment must be restarted from
$\ket{\psi}$. For small $\Delta\tau$ one has $\alpha \approx 1$ and
$\beta \approx -c\Delta\tau$, so $P_{\mathrm{s}}\approx 1/2$; the
protocol is therefore efficient at small step sizes.

\Cref{fig:lcu_circuit} summarises the LCU circuit graphically.

\begin{figure}[htbp]
\centering
\begin{quantikz}[column sep=10pt]
  \lstick[1]{anc.\ $\ket{0}$} & \gate{R_y(\vartheta)} & \ctrl{1}     & \gate{H}    & \meter{} \\
  \lstick[1]{$\ket{\psi}$}    & \qw                   & \gate{h}     & \qw         & \rstick{$\frac{(\alpha I + \beta h)\ket{\psi}}{\bigl\|(\alpha I + \beta h)\ket{\psi}\bigr\|}$ on $\ket{0}_{\mathrm{anc}}$} \qw
\end{quantikz}
\caption{LCU circuit implementing one truncated Trotter step
  $T^{(R)}(c\,\Delta\tau) = \alpha I + \beta h$ on the work register
  $\ket{\psi}$. The ancilla is rotated by an angle
  $\vartheta = 2\atantwo(\beta,\alpha)$, a controlled application of the
  Pauli string $h$ entangles ancilla and work, a Hadamard interferes the
  two LCU branches, and a measurement on the ancilla post-selects the
  desired branch. The work register collapses to the (renormalised)
  truncated step exactly when the measurement outcome is $0$, an event
  of probability $P_{\mathrm{s}}$ given by \cref{eq:lcu_success}.}
\label{fig:lcu_circuit}
\end{figure}

\subsection{The full TTITE algorithm}
\label{ssec:ttite_full}

Composing the building blocks of
\cref{ssec:trotter,ssec:taylor_step,ssec:lcu} yields the complete TTITE
procedure, summarised in \cref{alg:ttite}.

\begin{figure}[htbp]
\centering
\fbox{\begin{minipage}{0.94\textwidth}\small
\textbf{Algorithm: TTITE.} \\[2pt]
\textit{Input:} Pauli decomposition $H = \sum_{i=1}^{m} c_i\,h_i$;
initial state $\ket{\psi_0}$; total time $\tau$; step size $\Delta\tau$;
Taylor order $R$. \\
\textit{Output:} an approximation to
$e^{-H\tau}\ket{\psi_0}/\|e^{-H\tau}\ket{\psi_0}\|$.

\begin{enumerate}\setlength{\itemsep}{2pt}
  \item Set $L = \lceil\tau/\Delta\tau\rceil$ and
    $\ket{\psi}\leftarrow\ket{\psi_0}$.
  \item Repeat $L$ times:
  \begin{enumerate}
    \item For $i = 1,\ldots,m$:
    \begin{enumerate}
      \item Compute the Taylor coefficients $\alpha,\beta$ from
        \cref{eq:taylor_truncation} with $c=c_i$ and $h=h_i$.
      \item Build the LCU circuit of \cref{fig:lcu_circuit} with angle
        \cref{eq:lcu_angle} and controlled-Pauli string $h_i$.
      \item Execute the circuit on
        $\ket{0}_{\mathrm{anc}}\otimes\ket{\psi}$; on outcome $0$ update
        $\ket{\psi}$ with the (renormalised) post-selected state; on
        outcome $1$ \emph{restart} the current Trotter step.
    \end{enumerate}
  \end{enumerate}
  \item Return $\ket{\psi}$.
\end{enumerate}
\end{minipage}}
\caption{Pseudocode for the TTITE algorithm. Each Trotter step requires
  $m$ successful LCU post-selections; the per-segment success
  probability is given by \cref{eq:segment_success}.}
\label{alg:ttite}
\end{figure}

The algorithm is generic in the Hamiltonian: it uses only the Pauli
decomposition produced via \cref{eq:pauli_sum} and the (small) number of
qubits required by the encoding. In this work we apply it to the channel
Hamiltonians in the direct encoding of \cref{ssec:direct_encoding} (two
work qubits plus one ancilla, i.e.\ three qubits in total).

\subsection{Success probability and resource cost}
\label{ssec:ttite_resources}

Each Trotter step requires $m$ LCU circuits to succeed (one per Pauli
term), so the segment-level success probability factorises:
\begin{equation}
  \label{eq:segment_success}
  P_{\mathrm{seg}} = \prod_{i=1}^{m} P_{\mathrm{s},i},
\end{equation}
with $P_{\mathrm{s},i}$ given by \cref{eq:lcu_success}. For small
$\Delta\tau$ each individual factor approaches $1/2$, so the per-segment
probability decays exponentially in the number of Pauli terms. Two
observations soften this otherwise daunting cost.

First, Yi et al.~\cite{yi2025} prove that the success probability is
\emph{monotonically non-decreasing} in the imaginary-time step: as the
state approaches the ground state, applying $H - E_0$ becomes
quasi-tangential to it, which inflates the success probabilities
$P_{\mathrm{s},i}$. Concretely, with $\tau_\ell = \ell\,\Delta\tau$
denoting the cumulative imaginary time after the $\ell$-th segment,
\begin{equation}
  \label{eq:monotone_success}
  P_{\mathrm{seg}}(\tau_1) \;\leq\;
  P_{\mathrm{seg}}(\tau_2) \;\leq\;
  \cdots \;\leq\;
  P_{\mathrm{seg}}(\tau_L),
\end{equation}
so the algorithm becomes \emph{more} efficient as it converges.

Second, the failure probability $1 - P_{\mathrm{s},i}$ can be amplified
toward zero by amplitude amplification, at a cost of additional ancillas
and depth. We do not employ amplitude amplification in this work,
instead relying on the simpler repeat-until-success strategy described
in \cref{alg:ttite}. For the two-qubit channel Hamiltonians considered
here, the per-segment cost is at most a small multiplicative overhead,
and the algorithm remains practical for the imaginary times required to
reach the ground state.

The overall picture is therefore the following. TTITE bypasses the
parameter-space optimisation of VQE and VQITE by implementing the
imaginary-time propagator more or less directly on the device, at the
price of an ancilla and a post-selection step per Pauli term. Its
accuracy is controlled by two parameters, $\Delta\tau$ and $R$, and is
unaffected by ansatz expressibility.


% ──────────────────────────────────────────────────────────────────
\section{Summary}
\label{sec:framework-summary}

The framework developed in this chapter equips us with everything needed
to reproduce the three quantum simulations of charmonium spectroscopy
that this thesis benchmarks. \Cref{sec:common} produces three
$4\times 4$ real-symmetric channel matrices for ${}^1S_0$, ${}^3S_1$,
and ${}^1P_1$, together with the general procedure for expanding such a
matrix in Pauli strings. The Gallimore--Liao section
(\cref{sec:gallimore}) encodes the Hamiltonian fermionically through the
Jordan--Wigner transformation, uses a UCC ansatz, and minimises the
energy by classical optimisation; we extend their construction to the
spin-dependent $N=4$ case for comparability. The Woloshyn section
(\cref{sec:woloshyn}) encodes the Hamiltonian directly on two qubits,
uses a three-parameter $R_y$ ansatz, evolves it under the McLachlan
imaginary-time equations of motion, and computes M1 and E1 transition
amplitudes. The Yi et al.\ section (\cref{sec:yi}) skips the ansatz
entirely, implementing the imaginary-time propagator probabilistically
through LCU circuits on three qubits (two work, one ancilla).

\Cref{tab:algo_compare} collects these differences side by side. The
classical and quantum workloads listed are per iteration or per Trotter
step.

\begin{table}[htbp]
\centering
\caption{Side-by-side comparison of the three algorithms reproduced in
  this chapter, as applied to the channel Hamiltonians of
  \cref{eq:H_1S0,eq:H_3S1,eq:H_1P1}. Each method uses the encoding chosen
  by its source paper, with the small caveat that Gallimore and Liao's
  $N=3$ spinless setting is extended to $N=4$ with spin in
  \cref{ssec:gallimore_extension}.}
\label{tab:algo_compare}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{p{0.22\textwidth}p{0.24\textwidth}p{0.24\textwidth}p{0.24\textwidth}}
\toprule
                 & Gallimore--Liao (VQE)         & Woloshyn (VQITE)               & Yi et al.\ (TTITE)             \\
\midrule
Encoding         & Jordan--Wigner                 & Direct (compact)               & Direct (compact)               \\[2pt]
Qubits (work + anc.) & $N$ + 0                    & $\lceil\log_2 N\rceil$ + 0     & $\lceil\log_2 N\rceil$ + 1     \\[2pt]
Pauli terms (N=4) & $\sim 17$                     & $\sim 10$                      & $\sim 10$                      \\[2pt]
Ansatz           & UCC, $N{-}1$ params            & $R_y$ + CNOT, $3$ params       & none                            \\[2pt]
What is updated  & ansatz parameters              & ansatz parameters              & work-register state itself     \\[2pt]
What is optimised& energy expectation value       & McLachlan residual norm        & --- (no parameters)            \\[2pt]
Classical workload & gradient-free optimiser      & small linear-system solves     & track ancilla outcomes         \\[2pt]
Excited states     & orthogonality constraints    & state-overlap penalties        & not directly supported         \\[2pt]
Dominant error     & ansatz expressibility        & ansatz + Trotter-like step     & Trotter + Taylor + finite shots\\
\bottomrule
\end{tabular}
\end{table}

A few qualitative consequences of \cref{tab:algo_compare} deserve
emphasis. First, the three papers make two distinct encoding choices: the
fermionic (JW) encoding of Gallimore and Liao uses one qubit per HO basis
state, whereas the direct encoding of Woloshyn and our application of Yi
et al.\ uses one qubit per pair of basis states. The JW encoding
introduces a particle-conservation constraint on the ansatz, while the
direct encoding does not. Second, Gallimore--Liao and Woloshyn both
optimise parameters of a small ansatz manifold, so their accuracy is
bounded above by that manifold's expressibility; Yi et al.'s scheme
sidesteps the ansatz at the price of one ancilla qubit and a
post-selection per Pauli term. Third, transitions
(\cref{ssec:transitions}) are post-processing operations on eigenstates
produced by any of the three methods; their accuracy is therefore
directly inherited from the underlying eigenstate accuracy.

The next chapter applies all three algorithms to the channel
Hamiltonians of \cref{ssec:matrices}, quantifying these trade-offs
through numerical experiments with the codebase developed alongside this
thesis.
