\documentclass[12pt,a4paper]{report}

% ── Encoding and fonts ──
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}

% ── Page layout ──
\usepackage[margin=2.5cm]{geometry}
\usepackage{setspace}
\onehalfspacing

% ── Mathematics ──
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{bm}

% ── Graphics and diagrams ──
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning, shapes.geometric, calc}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usepackage{quantikz}

% ── Tables ──
\usepackage{booktabs}

% ── Captions and references ──
\usepackage[font=small,labelfont=bf]{caption}
\usepackage[hidelinks]{hyperref}
\usepackage[nameinlink,capitalize]{cleveref}

% ── Bibliography ──
\usepackage[backend=biber,style=numeric-comp,sorting=none]{biblatex}
\addbibresource{references.bib}

% ── Custom commands ──
% \ket, \bra, \braket provided by braket package (loaded via quantikz)
\newcommand{\expval}[1]{\left\langle#1\right\rangle}

% ══════════════════════════════════════════════════════════════════
\begin{document}

% ── Title page ──
\begin{titlepage}
\centering
\vspace*{2cm}
{\LARGE\bfseries Quantum Simulation of Spin-Dependent\\[6pt]
Charmonium Spectroscopy with\\[6pt]
Variational Quantum Algorithms\par}
\vspace{2cm}
{\Large Master's Thesis\par}
\vspace{1.5cm}
{\large Lucas Froguel\par}
\vfill
{\large\the\year\par}
\end{titlepage}

% ── Front matter ──
\pagenumbering{roman}
\tableofcontents
\clearpage
\pagenumbering{arabic}

% ══════════════════════════════════════════════════════════════════
\chapter{Introduction}
\label{ch:introduction}

The simulation of quantum systems lies at the heart of quantum computing's
promise. As Feynman observed in 1982, classical computers face a fundamental
barrier when modelling quantum mechanical systems: the Hilbert space of an
$n$-particle system grows exponentially, rendering exact classical computation
intractable for even modest system sizes~\cite{feynman1982}. Quantum computers,
by contrast, naturally encode and manipulate quantum states, offering a path
toward efficient simulation of physical systems that are otherwise
inaccessible~\cite{nielsen2010}.

This thesis applies variational quantum algorithms to the spectroscopy of
charmonium---bound states of a charm quark and its antiquark---including
spin-dependent interactions that produce the experimentally observed mass
splittings between states such as the $\eta_c$ and $J/\psi$ mesons. The
quarkonium system is physically rich enough to probe the capabilities of
near-term quantum algorithms across multiple quantum number channels and excited
states, yet small enough to be tractable with only a few qubits.

The remainder of this chapter introduces the three pillars on which this work
rests: the fundamentals of quantum computing (\cref{sec:qc}), variational
quantum methods (\cref{sec:variational}), and the quarkonium problem
(\cref{sec:quarkonium}).


% ──────────────────────────────────────────────────────────────────
\section{Quantum Computing}
\label{sec:qc}

\subsection{Why quantum computing?}
\label{sec:why_qc}

The classical theory of computation, formalized by Turing and Church in the
1930s, rests on the assumption that information is processed according to
classical physics. For decades this abstraction posed no practical limitation.
However, the simulation of quantum mechanical systems reveals a fundamental
barrier: the state space of an $n$-particle quantum system is a Hilbert space of
dimension $2^n$, so that a faithful classical representation requires resources
that grow exponentially with system size. A modest system of 50 interacting
spin-$\tfrac{1}{2}$ particles, for instance, demands $2^{50}\approx 10^{15}$
complex amplitudes---exceeding the memory of any existing classical computer.

In 1982, Feynman proposed a radical alternative: rather than simulating quantum
mechanics on a classical machine, one should use a controllable quantum system to
simulate another~\cite{feynman1982}. The exponential scaling that makes classical
simulation intractable becomes an \emph{asset}: a quantum processor with $n$
qubits naturally inhabits the same $2^n$-dimensional space as the target system.
This observation laid the conceptual foundation for quantum computing.

The subsequent development of quantum algorithms by Deutsch, Shor, and
Grover~\cite{nielsen2010} demonstrated that quantum computers are not merely
faster classical machines, but represent a genuinely different model of
computation. The complexity class $\mathsf{BQP}$ (bounded-error quantum
polynomial time) is believed to be strictly larger than its classical counterpart
$\mathsf{BPP}$, meaning there exist problems for which quantum algorithms
achieve an exponential advantage that no classical algorithm can
match~\cite{nielsen2010}. The source of this advantage lies in two uniquely
quantum phenomena: \emph{superposition}, which allows exploration of exponentially
many computational paths simultaneously, and \emph{interference}, which enables
the selective amplification of correct answers and suppression of incorrect ones.


\subsection{Qubits and quantum states}
\label{sec:qubits}

The fundamental unit of quantum information is the \emph{qubit}, a two-level
quantum system whose pure state is described by a normalized vector in
$\mathbb{C}^2$:
\begin{equation}
  \label{eq:qubit}
  \ket{\psi} = \alpha\ket{0} + \beta\ket{1},
  \qquad |\alpha|^2 + |\beta|^2 = 1,
\end{equation}
where $\alpha,\beta\in\mathbb{C}$ are \emph{probability amplitudes} and
$\{\ket{0},\ket{1}\}$ is the computational basis (conventionally identified with
the eigenstates of the Pauli $Z$ operator). Unlike a classical bit, which is
deterministically~0 or~1, a qubit exists in a coherent \emph{superposition} of
both basis states.

Geometrically, any single-qubit pure state can be represented as a point on the
\emph{Bloch sphere} via the parameterization
\begin{equation}
  \label{eq:bloch}
  \ket{\psi}
    = \cos\frac{\theta}{2}\,\ket{0}
    + e^{i\phi}\sin\frac{\theta}{2}\,\ket{1},
\end{equation}
with polar angle $\theta\in[0,\pi]$ and azimuthal angle $\phi\in[0,2\pi)$. The
north and south poles correspond to $\ket{0}$ and $\ket{1}$, respectively, while
points on the equator represent equal superpositions with varying relative phase.

For a system of $n$ qubits, the state space is the tensor product
$\mathcal{H} = (\mathbb{C}^2)^{\otimes n}$, spanned by $2^n$ computational
basis states $\ket{x_1 x_2 \cdots x_n}$ with $x_i\in\{0,1\}$. A general pure
state is therefore
\begin{equation}
  \label{eq:nqubit}
  \ket{\Psi} = \sum_{x\in\{0,1\}^n} c_x \ket{x},
  \qquad \sum_x |c_x|^2 = 1,
\end{equation}
requiring $2^n$ complex amplitudes to specify. It is this exponential growth of
the state space that makes quantum systems both hard to simulate classically and
powerful as computational devices.


\subsection{Entanglement}
\label{sec:entanglement}

A multi-qubit state is called \emph{separable} (or a product state) if it can be
written as a tensor product of individual qubit states,
$\ket{\Psi} = \ket{\psi_1}\otimes\ket{\psi_2}\otimes\cdots\otimes\ket{\psi_n}$.
States that \emph{cannot} be decomposed in this way are \emph{entangled}. The
simplest example is the Bell state
\begin{equation}
  \label{eq:bell}
  \ket{\Phi^+} = \frac{1}{\sqrt{2}}\bigl(\ket{00}+\ket{11}\bigr),
\end{equation}
which describes two qubits in a joint superposition: neither qubit has a definite
value, yet their measurement outcomes are perfectly correlated. This correlation
persists regardless of the spatial separation between the qubits and cannot be
explained by any local hidden variable theory, as demonstrated by Bell's
theorem~\cite{bell1964}.

Entanglement is not merely a theoretical curiosity---it is a computational
resource. Classical correlations can be shared between bits using shared
randomness, but entangled qubits carry correlations that are strictly stronger.
In quantum algorithms, entangling gates create superpositions over correlated
multi-qubit configurations, enabling the exploration of an exponentially large
solution space in a way that has no classical counterpart. Without entanglement, a
quantum computer can be efficiently simulated classically~\cite{nielsen2010};
it is entanglement that unlocks the full power of quantum computation.


\subsection{Quantum gates and circuits}
\label{sec:gates}

The time evolution of a closed quantum system is governed by a unitary operator.
Quantum computation exploits this by decomposing the desired evolution into a
sequence of elementary \emph{quantum gates}: unitary transformations acting on one
or two qubits at a time.

The three Pauli operators form a basis for single-qubit operations:
\begin{equation}
  \label{eq:pauli}
  X = \begin{pmatrix} 0 & 1 \\ 1 & 0 \end{pmatrix},\quad
  Y = \begin{pmatrix} 0 & -i \\ i & 0 \end{pmatrix},\quad
  Z = \begin{pmatrix} 1 & 0 \\ 0 & -1 \end{pmatrix}.
\end{equation}
Together with the identity $I$, they satisfy $\sigma_i^2 = I$, anticommute
pairwise ($\sigma_i\sigma_j = -\sigma_j\sigma_i$ for $i\neq j$), and form a
basis for the space of $2\times 2$ Hermitian matrices. Any $2\times 2$ Hermitian
operator---and hence any single-qubit observable---can be written as
$A = a_0 I + a_1 X + a_2 Y + a_3 Z$ with $a_i\in\mathbb{R}$.

The Pauli operators generate continuous families of rotation gates on the Bloch
sphere:
\begin{equation}
  \label{eq:rotations}
  R_\mu(\theta) = e^{-i\theta\,\mu/2}
    = \cos\frac{\theta}{2}\,I - i\sin\frac{\theta}{2}\,\mu,
  \qquad \mu \in \{X,Y,Z\}.
\end{equation}
The Hadamard gate $H = (X+Z)/\sqrt{2}$ maps $\ket{0}\mapsto(\ket{0}+\ket{1})/\sqrt{2}$,
creating an equal superposition from a computational basis state.

Two-qubit entangling gates introduce correlations between qubits. The
controlled-NOT (CNOT) gate flips the target qubit conditioned on the control
qubit being in state $\ket{1}$:
\begin{equation}
  \label{eq:cnot}
  \text{CNOT} = \ket{0}\bra{0}\otimes I + \ket{1}\bra{1}\otimes X
  = \begin{pmatrix}
    1 & 0 & 0 & 0 \\
    0 & 1 & 0 & 0 \\
    0 & 0 & 0 & 1 \\
    0 & 0 & 1 & 0
  \end{pmatrix}.
\end{equation}
The set $\{R_X,R_Y,R_Z,\text{CNOT}\}$ constitutes a \emph{universal gate set}:
any unitary on $n$ qubits can be approximated to arbitrary precision by a finite
sequence of these gates~\cite{nielsen2010}.

A quantum algorithm is specified as a \emph{quantum circuit}: a sequence of gates
applied to a register of qubits initialized in a known state, followed by
measurement. \Cref{fig:bell_circuit} illustrates a minimal circuit that prepares
the Bell state of \cref{eq:bell} using only a Hadamard gate and a CNOT.

\begin{figure}[htbp]
\centering
\begin{quantikz}
  \lstick{$\ket{0}$} & \gate{H} & \ctrl{1} & \meter{} \\
  \lstick{$\ket{0}$} & \qw      & \targ{}  & \meter{}
\end{quantikz}
\caption{Quantum circuit preparing the Bell state
  $\ket{\Phi^+}=(\ket{00}+\ket{11})/\sqrt{2}$. The Hadamard gate creates a
  superposition on the first qubit; the CNOT entangles both qubits. Measurement
  yields $\ket{00}$ or $\ket{11}$, each with probability $1/2$.}
\label{fig:bell_circuit}
\end{figure}


\subsection{Quantum parallelism and interference}
\label{sec:parallelism}

The mechanism behind quantum computational advantage is often described loosely
as ``computing on all inputs at once.'' While evocative, this picture is
incomplete: the key ingredient is not parallelism alone, but the ability to make
the parallel branches \emph{interfere}.

Consider a unitary $U_f$ that evaluates a function $f:\{0,1\}^n\to\{0,1\}$.
Applied to an equal superposition over all $2^n$ inputs, it produces
\begin{equation}
  \label{eq:parallelism}
  U_f \frac{1}{\sqrt{2^n}}\sum_{x=0}^{2^n-1}\ket{x}\ket{0}
  = \frac{1}{\sqrt{2^n}}\sum_{x=0}^{2^n-1}\ket{x}\ket{f(x)}.
\end{equation}
All function values are encoded in a single quantum state. However, measuring
this state yields a \emph{single} random pair $(x,f(x))$---no better than
classical evaluation. The power of quantum algorithms lies in subsequent
operations that exploit \emph{interference}: unitary gates are designed so that
amplitudes corresponding to desired outputs add constructively, while those
corresponding to undesired outputs cancel destructively. The final measurement
then yields the correct answer with high probability.

This interplay of superposition, entanglement, and interference is the
mathematical core of quantum speedups. Superposition provides exponential
parallelism; entanglement enables correlations across qubits that have no
classical equivalent; and interference channels the computation toward the
correct outcome.


\subsection{Measurement and the Born rule}
\label{sec:measurement}

While unitary evolution is deterministic, extracting information from a quantum
system is inherently probabilistic. A \emph{projective measurement} in the
computational basis is described by the set of projectors
$\{\Pi_x = \ket{x}\bra{x}\}_{x\in\{0,1\}^n}$. Given a state
$\ket{\Psi}=\sum_x c_x\ket{x}$, the \emph{Born rule} prescribes the outcome
probabilities:
\begin{equation}
  \label{eq:born}
  p(x) = |\braket{x}{\Psi}|^2 = |c_x|^2.
\end{equation}
Upon obtaining outcome $x$, the state collapses to $\ket{x}$, and any
information encoded in the remaining amplitudes is irretrievably lost.

This has a direct practical consequence: a single measurement of $n$ qubits
yields only $n$ classical bits, despite the state containing $2^n$ amplitudes.
To estimate a continuous quantity---such as an energy expectation value---the
measurement must be repeated many times ($N_{\text{shots}}$ repetitions), and the
result is subject to statistical uncertainty scaling as $1/\sqrt{N_{\text{shots}}}$.

More generally, to measure an observable $O$ (a Hermitian operator), its
expectation value in state $\ket{\Psi}$ is given by
\begin{equation}
  \label{eq:expectation}
  \expval{O} = \bra{\Psi}O\ket{\Psi} = \sum_i \lambda_i\, p_i,
\end{equation}
where $\lambda_i$ are the eigenvalues of $O$ and $p_i$ the probability of
obtaining the corresponding eigenstate. When $O$ is decomposed as a sum of Pauli
operators, each Pauli term can be measured independently by rotating to the
appropriate eigenbasis before measurement in the computational basis.


\subsection{Density matrices and mixed states}
\label{sec:density}

The pure state formalism of \cref{eq:qubit} describes isolated, perfectly
controlled quantum systems. Real quantum processors, however, interact with their
environment, leading to \emph{decoherence} and \emph{mixed states} that require a
more general description.

The \emph{density matrix} (or density operator) of a pure state
$\ket{\psi}$ is
\begin{equation}
  \label{eq:density_pure}
  \rho = \ket{\psi}\bra{\psi}.
\end{equation}
A \emph{mixed state}---representing classical uncertainty about which pure state
the system is in---is described by a convex combination
\begin{equation}
  \label{eq:density_mixed}
  \rho = \sum_i p_i \ket{\psi_i}\bra{\psi_i},
  \qquad p_i \geq 0,\quad \sum_i p_i = 1.
\end{equation}
The density matrix is Hermitian, positive semidefinite, and has unit trace. A
state is pure if and only if $\mathrm{Tr}(\rho^2) = 1$; for mixed states,
$\mathrm{Tr}(\rho^2) < 1$, quantifying the degree of decoherence.

In the density matrix formalism, unitary evolution becomes
$\rho \mapsto U\rho\, U^\dagger$, and expectation values are computed as
$\expval{O}=\mathrm{Tr}(O\rho)$. Noise processes---such as the depolarizing
channel relevant to this thesis---are described by completely positive
trace-preserving (CPTP) maps on $\rho$. For example, the single-qubit
depolarizing channel with error rate $p$ acts as
\begin{equation}
  \label{eq:depolarizing}
  \mathcal{E}(\rho) = (1-p)\,\rho + \frac{p}{3}
    \bigl(X\rho X + Y\rho Y + Z\rho Z\bigr),
\end{equation}
which with probability $p$ replaces the state with a random Pauli error.
This formalism will be central to the analysis of noise effects in
\cref{ch:introduction} and subsequent chapters.


\subsection{Quantum simulation}
\label{sec:qsim}

Among the many proposed applications of quantum computing---including
cryptanalysis, optimization, and machine learning---the simulation of quantum
systems stands out as the most natural and arguably the most promising near-term
application~\cite{feynman1982,preskill2018}. The reason is both conceptual and
practical: a quantum computer \emph{is} a quantum system, so using it to simulate
another quantum system leverages its native capabilities rather than working
against them.

The essential mapping is straightforward. A physical system with a
$2^n$-dimensional Hilbert space is encoded on $n$ qubits. The system's
Hamiltonian $H$ is decomposed into a sum of Pauli operators acting on these
qubits:
\begin{equation}
  \label{eq:pauli_ham}
  H = \sum_{i=1}^{M} c_i\, P_i,
  \qquad P_i \in \{I,X,Y,Z\}^{\otimes n},
\end{equation}
where each $P_i$ is a tensor product of single-qubit Pauli matrices (a
\emph{Pauli string}) and $c_i\in\mathbb{R}$ are real coefficients. This
decomposition is always possible since the $4^n$ Pauli strings form a complete
basis for $2^n\times 2^n$ Hermitian matrices.

Ground state energies, excited state spectra, and dynamical properties can then
be extracted using quantum algorithms that prepare and measure appropriate states
on the quantum register. For the small systems considered in this thesis, the
encoding requires only 2--3 qubits, but the same framework scales to larger
systems where classical methods become intractable.

Quantum simulation of atomic, molecular, and nuclear systems has seen rapid
progress in recent years, with demonstrations ranging from small molecules to
lattice gauge theories~\cite{peruzzo2014}. The quarkonium system studied in this
thesis belongs to this broader programme, representing a step toward quantum
simulation of QCD bound states.


\subsection{The NISQ era}
\label{sec:nisq}

Current quantum hardware operates in the \emph{Noisy Intermediate-Scale Quantum}
(NISQ) regime~\cite{preskill2018}, characterized by processors with tens to
hundreds of qubits subject to significant imperfections:
\begin{itemize}
  \item \textbf{Gate errors} on the order of $10^{-3}$--$10^{-2}$ per operation,
    described by noise channels such as the depolarizing channel of
    \cref{eq:depolarizing};
  \item \textbf{Limited connectivity} between qubits, requiring additional SWAP
    gates that increase circuit depth;
  \item \textbf{Short coherence times} that bound the number of operations
    executable before decoherence degrades the quantum state.
\end{itemize}
These constraints preclude the deep circuits required by fault-tolerant algorithms
such as quantum phase estimation, which need thousands of controlled operations
and ancilla qubits for error correction.

The NISQ era has therefore driven the development of algorithms specifically
designed for shallow circuits and resilient to moderate noise levels. The most
successful paradigm to date is the \emph{variational} approach, in which a
quantum processor prepares parameterized trial states while a classical computer
optimizes the parameters---a hybrid strategy discussed in the next section.


% ──────────────────────────────────────────────────────────────────
\section{Variational Quantum Methods}
\label{sec:variational}

\subsection{The variational principle}

Variational methods offer a framework for extracting useful results from NISQ
hardware. The central idea rests on the \emph{Rayleigh--Ritz variational
principle}: for a Hamiltonian $H$ with ground state energy $E_0$,
\begin{equation}
  \label{eq:variational}
  E_0 \leq \bra{\psi(\bm{\theta})}H\ket{\psi(\bm{\theta})}
       \equiv E(\bm{\theta})
\end{equation}
for any normalized trial state $\ket{\psi(\bm{\theta})}$ parameterized by
classical variables $\bm{\theta}\in\mathbb{R}^p$. Minimizing
$E(\bm{\theta})$ over the parameter space yields an upper bound on $E_0$, with
equality when $\ket{\psi(\bm{\theta})}$ includes the true ground state in its
span.


\subsection{The hybrid quantum--classical loop}

In the quantum computing implementation~\cite{peruzzo2014,mcclean2016}, the
trial state is prepared by a \emph{parameterized quantum circuit} (also called an
\emph{ansatz}) applied to an initial reference state:
\begin{equation}
  \label{eq:ansatz}
  \ket{\psi(\bm{\theta})}
    = U(\bm{\theta})\ket{0}^{\otimes n},
\end{equation}
where $U(\bm{\theta})$ is a product of parameterized and fixed gates. The energy
expectation value is then estimated on the quantum processor. For a Hamiltonian
decomposed as a weighted sum of Pauli strings,
\begin{equation}
  \label{eq:pauli_decomp}
  H = \sum_{i=1}^{M} c_i\, P_i,
  \qquad P_i \in \{I,X,Y,Z\}^{\otimes n},
\end{equation}
each expectation value $\expval{P_i}$ is estimated independently through
repeated state preparation and measurement in the appropriate Pauli basis. The
total energy,
\begin{equation}
  \label{eq:energy_estimator}
  E(\bm{\theta}) = \sum_{i=1}^{M} c_i \expval{P_i}_{\bm{\theta}},
\end{equation}
is passed to a \emph{classical optimizer} that updates $\bm{\theta}$ to
minimize~$E$. This loop repeats until convergence, as illustrated in
\cref{fig:vqe_loop}.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}[
    node distance=1.6cm and 2.5cm,
    box/.style={
      draw, rounded corners=4pt, minimum width=3.2cm,
      minimum height=1cm, align=center, font=\small
    },
    arrow/.style={-{Stealth[length=6pt]}, thick},
  ]
  % Nodes
  \node[box, fill=blue!8] (prep) {Prepare\\$\ket{\psi(\bm{\theta})}$};
  \node[box, fill=blue!8, right=of prep] (meas)
    {Measure\\$\expval{P_i}_{\bm{\theta}}$};
  \node[box, fill=orange!10, below=of meas] (eval)
    {Compute\\$E(\bm{\theta})=\sum c_i\expval{P_i}$};
  \node[box, fill=orange!10, below=of prep] (opt)
    {Classical optimizer\\update $\bm{\theta}$};

  % Arrows
  \draw[arrow] (prep) -- (meas);
  \draw[arrow] (meas) -- (eval);
  \draw[arrow] (eval) -- (opt);
  \draw[arrow] (opt) -- (prep);

  % Labels
  \node[font=\footnotesize\itshape, above=0.2cm of prep, anchor=south west,
        xshift=-0.3cm] {Quantum processor};
  \node[font=\footnotesize\itshape, below=0.2cm of opt, anchor=north west,
        xshift=-0.3cm] {Classical computer};

  % Dashed separator
  \draw[dashed, gray] ($(prep.south west)+(-0.7,-0.8)$) --
                       ($(meas.south east)+(0.7,-0.8)$);
\end{tikzpicture}
\caption{Schematic of the hybrid quantum--classical variational loop. The
  quantum processor prepares a parameterized state and estimates Pauli
  expectation values; the classical computer evaluates the energy and updates
  circuit parameters. The loop iterates until convergence.}
\label{fig:vqe_loop}
\end{figure}


\subsection{Ansatz design and practical considerations}

The choice of ansatz circuit is problem-dependent and critical to performance.
\emph{Hardware-efficient} ansatze use gates native to the target device,
minimizing circuit depth at the cost of physical motivation.
\emph{Physically-motivated} ansatze, such as those inspired by unitary coupled
cluster theory from quantum chemistry, encode domain knowledge into the circuit
structure and may converge faster toward the physical ground state.

The hybrid approach offers several advantages for near-term hardware. The quantum
processor handles the exponentially hard task of preparing and measuring
many-body quantum states, while the classical optimizer navigates the energy
landscape using well-established numerical methods. Circuit depths remain
manageable, and variational methods exhibit a degree of inherent noise
resilience: small perturbations in the energy surface due to hardware errors are
partially absorbed by the optimization~\cite{mcclean2016}.

Beyond the ground state, variational frameworks extend naturally to excited
states through orthogonality constraints and penalty methods, and to quantum
dynamics through variational formulations of time evolution. These extensions
broaden the scope of problems addressable on near-term quantum hardware.


% ──────────────────────────────────────────────────────────────────
\section{The Quarkonium Problem}
\label{sec:quarkonium}

\subsection{Heavy quark bound states}

Quantum chromodynamics (QCD), the gauge theory of the strong interaction,
governs the binding of quarks and gluons into hadrons. Among the simplest
hadronic systems are \emph{quarkonia}: bound states of a heavy quark and its
antiquark ($q\bar{q}$). Charmonium ($c\bar{c}$) systems have served as testing
grounds for QCD since the simultaneous discovery of the $J/\psi$ meson in 1974
at BNL~\cite{aubert1974} and SLAC~\cite{augustin1974}---an event known as the
\emph{November Revolution} in particle physics.

The large mass of the charm quark ($m_c \approx 1.27$~GeV/$c^2$) implies that
the quark--antiquark pair moves non-relativistically inside the bound state, with
typical velocities $v/c \sim 0.3$. This justifies a potential model
approach~\cite{eichten1978} in which the dynamics are governed by the
non-relativistic Schr\"{o}dinger equation with an effective interquark potential.


\subsection{The Cornell potential}

The \emph{Cornell potential}~\cite{eichten1978},
\begin{equation}
  \label{eq:cornell}
  V(r) = -\frac{\kappa}{r} + \sigma\, r,
\end{equation}
provides a successful phenomenological description of the $q\bar{q}$ interaction.
It captures the two essential features of the strong force:
\begin{itemize}
  \item \textbf{Short-range Coulomb attraction} ($-\kappa/r$), arising from
    one-gluon exchange between the color charges, analogous to the
    electromagnetic Coulomb potential but with QCD coupling;
  \item \textbf{Long-range linear confinement} ($\sigma\,r$), a consequence of
    the non-Abelian nature of QCD, modelled by a chromoelectric flux tube
    (string) with tension $\sigma$ connecting the quark and antiquark.
\end{itemize}
Typical parameter values for charmonium are $\kappa \approx 0.52$ and
$\sigma \approx 0.183$~GeV$^2$~\cite{eichten1978}. \Cref{fig:cornell_potential}
shows the characteristic shape of this potential.

\begin{figure}[htbp]
\centering
\begin{tikzpicture}
\begin{axis}[
    width=0.65\textwidth,
    height=0.45\textwidth,
    xlabel={$r$ (fm)},
    ylabel={$V(r)$ (GeV)},
    xmin=0.1, xmax=2.0,
    ymin=-2.5, ymax=2.0,
    domain=0.1:2.0,
    samples=200,
    grid=major,
    grid style={gray!30},
    legend pos=south east,
    legend style={font=\small},
    every axis label/.style={font=\small},
    tick label style={font=\footnotesize},
  ]
  % Coulomb part
  \addplot[blue, dashed, thick]
    {-0.52/x};
  \addlegendentry{$-\kappa/r$}
  % Linear part
  \addplot[red, dashed, thick]
    {0.183*x/0.197};
  \addlegendentry{$\sigma r$}
  % Cornell (total)
  \addplot[black, very thick]
    {-0.52/x + 0.183*x/0.197};
  \addlegendentry{$V(r) = -\kappa/r + \sigma r$}
\end{axis}
\end{tikzpicture}
\caption{The Cornell potential as a function of interquark separation $r$. The
  dashed curves show the Coulomb ($-\kappa/r$, blue) and linear confinement
  ($\sigma r$, red) components; the solid curve is their sum. At short distances
  the potential is dominated by one-gluon exchange; at large distances the
  linearly rising string tension ensures quark confinement.}
\label{fig:cornell_potential}
\end{figure}


\subsection{Spin-dependent interactions}

The spin-independent Cornell potential alone does not account for the full
charmonium spectrum. The mass splitting between states with different spin
quantum numbers---for instance, between the $\eta_c$ (${}^1S_0$, spin singlet)
and the $J/\psi$ (${}^3S_1$, spin triplet)---requires spin-dependent
corrections. At leading order, the dominant contribution for $S$-wave states is
the hyperfine contact interaction~\cite{brambilla2011},
\begin{equation}
  \label{eq:spin}
  V_{\text{hyp}}(r) = \frac{32\pi\,\alpha_s}{9\,m_c^2}\,
  \delta^{(3)}(\mathbf{r})\;\mathbf{S}_c\cdot\mathbf{S}_{\bar{c}},
\end{equation}
where $\alpha_s$ is the strong coupling constant and
$\mathbf{S}_c\cdot\mathbf{S}_{\bar{c}}$ is the spin--spin dot product. The
eigenvalues of the latter operator distinguish singlet and triplet
configurations:
\begin{equation}
  \label{eq:spin_eigenvalues}
  \mathbf{S}_c\cdot\mathbf{S}_{\bar{c}} =
  \begin{cases}
    -\tfrac{3}{4} & \text{spin singlet } (S=0),\\[4pt]
    +\tfrac{1}{4} & \text{spin triplet } (S=1).
  \end{cases}
\end{equation}

Quarkonium states are classified in spectroscopic notation ${}^{2S+1}L_J$, where
$S$ is the total spin, $L$ the orbital angular momentum ($S,P,D,\ldots$ for
$L=0,1,2,\ldots$), and $J$ the total angular momentum. The low-lying charmonium
states relevant to this work are:
\begin{itemize}
  \item ${}^1S_0$: spin singlet, $S$-wave $\longrightarrow$ $\eta_c$ meson
    ($M \approx 2984$~MeV),
  \item ${}^3S_1$: spin triplet, $S$-wave $\longrightarrow$ $J/\psi$ meson
    ($M \approx 3097$~MeV),
  \item ${}^1P_1$: spin singlet, $P$-wave $\longrightarrow$ $h_c$ meson
    ($M \approx 3526$~MeV).
\end{itemize}


\subsection{Quarkonium as a quantum simulation testbed}

The full quarkonium Hamiltonian, including kinetic energy and both
spin-independent and spin-dependent potentials, reads
\begin{equation}
  \label{eq:hamiltonian}
  H = -\frac{1}{2\mu}\nabla^2 + V(r) + V_{\text{hyp}}(r),
\end{equation}
where $\mu = m_c/2$ is the reduced mass. Expanding the radial wavefunction in a
truncated basis of $N$ quantum harmonic oscillator (QHO) eigenstates yields a
finite-dimensional matrix eigenvalue problem of size $N\times N$, whose entries
are determined by the matrix elements of the kinetic and potential operators in
the QHO basis.

This matrix can then be decomposed into a sum of Pauli operators
(\cref{eq:pauli_decomp}) and encoded on a quantum register of
$\lceil\log_2 N\rceil$ qubits, making it directly amenable to variational
quantum algorithms. With $N=3$--$4$ basis states, the problem requires only 2--3
qubits, placing it firmly within reach of current NISQ devices while retaining
the essential physics: multiple spectroscopic channels, spin-dependent
splittings, and a hierarchy of excited states.

This combination of physical richness and computational tractability makes
quarkonium an ideal testbed for benchmarking quantum simulation algorithms
against exact solutions and experimental data.


% ──────────────────────────────────────────────────────────────────
\section{Thesis Overview}
\label{sec:overview}

This thesis investigates the application of variational quantum algorithms to the
charmonium spectroscopy problem described above. We implement and compare several
algorithmic approaches for determining both ground and excited state energies
across the ${}^1S_0$, ${}^3S_1$, and ${}^1P_1$ channels, assess their
performance under realistic noise conditions, and evaluate error mitigation
strategies. The goal is twofold: to demonstrate the feasibility of quantum
simulation for hadronic physics on near-term hardware, and to provide a
systematic comparison of methods that may inform future applications to more
complex systems.

The thesis is organized as follows:
\begin{itemize}
  \item \textbf{Chapter~2} develops the theoretical framework, including the
    Hamiltonian construction in the harmonic oscillator basis and the mapping to
    qubit operators.
  \item \textbf{Chapter~3} presents the quantum algorithms employed, covering
    both eigenvalue-based and time-evolution-based variational approaches.
  \item \textbf{Chapter~4} details the numerical results: energy spectra,
    convergence properties, and cross-method comparisons.
  \item \textbf{Chapter~5} addresses the effects of hardware noise and
    discusses error mitigation techniques.
  \item \textbf{Chapter~6} concludes with a summary of findings and an outlook
    on future directions.
\end{itemize}


% ══════════════════════════════════════════════════════════════════
% Placeholder for future chapters
% \chapter{Theoretical Framework}
% \chapter{Quantum Algorithms}
% \chapter{Results}
% \chapter{Noise and Error Mitigation}
% \chapter{Conclusions and Outlook}

% ── Bibliography ──
\printbibliography[heading=bibintoc]

\end{document}
